---
title: |
  Hierarchical Bayesian Football Performance Model\
  Pipeline Results Journal
subtitle: La Liga — Low-Rank Player Effects + Diagonal Season Effects
author: "S. Serra Pe"
date: today
format:
  pdf:
    documentclass: article
    classoption: [11pt, a4paper]
    geometry: "margin=2.5cm"
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    linkcolor: "blue!60!black"
    urlcolor: "blue!60!black"
    pdf-engine: pdflatex
    include-in-header:
      text: |
        \usepackage{amsmath, amssymb}
        \usepackage{bm}
        \usepackage{booktabs}
        \usepackage{float}
        \usepackage{xcolor}
        \usepackage{graphicx}
        \usepackage{caption}
        \usepackage{subcaption}
        \usepackage{array}
        \usepackage{longtable}
        \floatplacement{figure}{H}
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(readr)
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)

# Resolve model root from the location of this .qmd file
model_root <- normalizePath(dirname(knitr::current_input(dir = TRUE)))

tables_dir      <- file.path(model_root, "outputs", "tables")
figures_dir     <- file.path(model_root, "outputs", "figures")
diagnostics_dir <- file.path(model_root, "outputs", "diagnostics")

# Helpers
read_t <- function(f) {
  p <- file.path(tables_dir, f)
  if (file.exists(p)) readr::read_csv(p, show_col_types = FALSE) else NULL
}
read_d <- function(f) {
  p <- file.path(diagnostics_dir, f)
  if (file.exists(p)) readr::read_csv(p, show_col_types = FALSE) else NULL
}
fig <- function(f) file.path(figures_dir, f)
chain_fig <- function(subdir, f) file.path(figures_dir, "11_chain_plots", subdir, f)

show_table <- function(df, caption = NULL, digits = 4) {
  if (is.null(df)) return(invisible(NULL))
  knitr::kable(df, caption = caption, digits = digits,
               booktabs = TRUE, linesep = "") |>
    kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"),
                              font_size = 9)
}
```

---

# Project Overview

## Model

The pipeline decomposes La Liga player-season performance using an additive hierarchical model:

$$\mathbf{y}_n = \boldsymbol{\mu} + \mathbf{a}_{i[n]} + \mathbf{b}_{j[n]} + \boldsymbol{\varepsilon}_n, \qquad n = 1, \ldots, N$$

where $\mathbf{y}_n \in \mathbb{R}^P$ is a vector of $P$ scaled features, $\mathbf{a}_i \sim \mathcal{N}(\mathbf{0}, \Sigma_a)$ is a player effect, $\mathbf{b}_j \sim \mathcal{N}(\mathbf{0}, \Sigma_b)$ is a season effect, and $\boldsymbol{\varepsilon}_n \sim \mathcal{N}(\mathbf{0}, \Sigma_\varepsilon)$ is residual noise. The key contribution is the **low-rank player + diagonal season** specification (stages 09/10):

$$\Sigma_a = \Lambda_a \Lambda_a^\top + \Psi_a, \qquad \Sigma_b = \mathrm{Diag}(\sigma_b^2)$$

The factor loadings $\Lambda_a \in \mathbb{R}^{P \times Q_a}$ capture latent player archetypes. The diagonal season structure is motivated by the small number of seasons ($S = 12$), which provides insufficient data to identify cross-feature seasonal covariance.

## Data

| Item | Value |
|---|---|
| Source | Wyscout La Liga event data |
| Seasons | 12 (2014/15 – 2025/26) |
| Sample | Non-GK players $\geq 450$ min/season |
| Player-season rows $N$ | 4,944 |
| Unique players $I$ | 1,650 |
| Features $P$ | 53 (scaled per-90 + success rates) |
| Player factor rank $Q_a$ | 2 (default; set via `BFA_SIM_RANK_A`) |

## Pipeline Stages

```{r pipeline-stages-table}
tibble::tribble(
  ~Stage, ~Script, ~Purpose,
  "00", "00_build_longitudinal_dataset.R",      "Aggregate match rows to player-season panel",
  "01", "01_read_clean_feature_engineering.R",  "Clean, filter, engineer 53 features",
  "02", "02_prepare_model_objects.R",            "Scale data, build indices and Stan data",
  "03", "03_fit_real_diagonal_additive.R",       "Baseline diagonal model — real data",
  "04", "04_simulate_lowrank_additive_data.R",  "Simulate data with known low-rank truth",
  "05", "05_fit_sim_diagonal_additive.R",        "Diagonal model — simulated data (misspecification check)",
  "06", "06_fit_sim_lowrank_recovery.R",         "Full low-rank recovery — failed (max Rhat = 1.74)",
  "07", "07_diagnostics_and_recovery_plots.R",  "Recovery tables and plots for stages 03/05/06",
  "08", "08_validate_all_models.R",              "Comprehensive validation (sampler, posterior, LOO)",
  "09", "09_fit_sim_lowrank_a_diag_b.R",         "New model — simulated data (recovery validation)",
  "10", "10_fit_real_lowrank_a_diag_b.R",        "New model — real La Liga data (main result)",
  "11", "11_loading_analysis_and_chains.R",      "Loading tables, chain plots, recovery metrics",
  "12", "12_football_interpretation.R",          "Factor interpretation, player archetypes"
) |>
  knitr::kable(booktabs = TRUE, linesep = "", caption = "Pipeline stage reference") |>
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9) |>
  kableExtra::column_spec(1, width = "0.8cm") |>
  kableExtra::column_spec(2, width = "5.5cm") |>
  kableExtra::column_spec(3, width = "7cm")
```

## Environment Variables

| Variable | Default | Description |
|---|---|---|
| `BFA_RUN_STAN` | `false` | Set `true` to run Stan sampling |
| `BFA_REUSE_FIT` | `true` | Reuse existing CSV fits if found |
| `BFA_CHAINS` | `4` | Number of MCMC chains |
| `BFA_PARALLEL_CHAINS` | `4` | Parallel chains |
| `BFA_ITER_WARMUP` | `1000` | Warmup iterations per chain |
| `BFA_ITER_SAMPLING` | `1000` | Sampling iterations per chain |
| `BFA_ADAPT_DELTA` | `0.95` | Target acceptance probability |
| `BFA_MAX_TREEDEPTH` | `12` | Max NUTS tree depth |
| `BFA_SIM_RANK_A` | `2` | Player factor rank $Q_a$ |
| `BFA_SEED` | `20260513` | Random seed |

---

# Journal Entries

---

## Entry 1 — 2026-05-27 — Stages 03–12: Baseline through New Low-Rank Model

### Configuration

| Parameter | Value |
|---|---|
| Seed | 20260513 |
| Chains | 4 |
| Parallel chains | 4 |
| Iter warmup | 1000 |
| Iter sampling | 1000 |
| Adapt delta | 0.95 |
| Max treedepth | 12 |
| Player factor rank $Q_a$ | 2 |

### Stages Run

- **03** — Diagonal additive model fitted to real La Liga data. Establishes baseline variance decomposition.
- **05** — Diagonal additive model fitted to simulated low-rank data. Validates that diagonal model recovers diagonal elements well even under misspecification.
- **06** — Full low-rank recovery (both $\Sigma_a$ and $\Sigma_b$ as factor models). Failed to converge: both $\Sigma_a$ and $\Sigma_b$ factor loadings poorly mixed. Diagnosed as identifiability problem with only 12 seasons.
- **07** — Recovery plots and diagnostic tables for stages 03/05/06.
- **08** — Comprehensive validation: sampler health, posterior health, PSIS-LOO. LOO estimates unreliable (Pareto $k \gg 1$ for most observations).
- **09** — New model (low-rank $\Sigma_a$ + diagonal $\Sigma_b$) fitted to simulated data. Validates Factor 2 recovery; Factor 1 still poorly mixed.
- **10** — New model fitted to real La Liga data. Main substantive result.
- **11** — Loading tables, chain plots, recovery metrics vs simulation truth. Two bugs fixed in this run (wrong truth field names, wrong matrix indexing for scatter plot).
- **12** — Football-level factor interpretation and player archetype labelling.

### Sampler Health

#### Stages 03, 05, 06 (from stage 08 validation)

```{r sampler-health-08}
sh <- read_d("08_sampler_health_summary.csv")
if (!is.null(sh)) {
  sh |>
    select(model_id, n_draws, n_divergent, max_treedepth,
           n_max_treedepth_hits, min_ebfmi, mean_accept_stat) |>
    mutate(
      model_id = sub("^0[0-9]+_", "", model_id),
      model_id = sub("_additive$", "", model_id),
      model_id = sub("_recovery$", "_recov", model_id),
      across(where(is.numeric), \(x) round(x, 4))
    ) |>
    rename(
      Model        = model_id,
      Draws        = n_draws,
      Diverg.      = n_divergent,
      MaxTree      = max_treedepth,
      TreeHits     = n_max_treedepth_hits,
      MinEBFMI     = min_ebfmi,
      AcceptStat   = mean_accept_stat
    ) |>
    knitr::kable(booktabs = TRUE, linesep = "",
                 caption = "Sampler health — stages 03/05/06") |>
    kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"),
                              font_size = 9) |>
    kableExtra::column_spec(1, width = "4cm")
}
```

#### Stages 09 and 10 (from stage 11 analysis)

```{r sampler-health-11}
ph <- read_d("11_posterior_health_new_model.csv")
if (!is.null(ph)) {
  ph |>
    mutate(across(where(is.numeric), \(x) round(x, 4))) |>
    show_table(caption = "Posterior health — stages 09/10 (low-rank $\\Sigma_a$ + diagonal $\\Sigma_b$)")
}
```

**Interpretation:** Stages 03/05/06 have clean geometry (zero divergences, no treedepth saturation). Stages 09/10 also have zero divergences but show high $\hat{R}$ ($\approx 1.75$–$1.78$) and very low ESS ($\approx 6$) for Factor 1 loading parameters. This is a **mixing failure**, not a geometry failure — the sampler accepts proposals but explores the loading space inefficiently due to sign/scale ambiguity in the first factor. Factor 2 and all variance components converge cleanly.

### Recovery Metrics (Stage 09 vs Simulation Truth)

```{r recovery-metrics}
rec <- read_t("11_sim_recovery_metrics.csv")
if (!is.null(rec)) {
  rec |>
    select(parameter, n, bias, mae, rmse, corr, cov90) |>
    mutate(across(where(is.numeric), \(x) round(x, 4))) |>
    arrange(parameter) |>
    show_table(caption = "Stage 09 recovery vs simulation truth (posterior mean vs true value)")
}
```

**Interpretation:**

- **$\Lambda_a$ Factor 2**: Near-perfect recovery (corr $= 0.999$, MAE $= 0.013$). Factor 2 is identified and mixing well.
- **$\Lambda_a$ Factor 1**: Corr $= 0.305$ — essentially unrecovered. The posterior median is weakly correlated with the truth, though coverage is high (96.2%) because the posterior is wide.
- **$\Sigma_a$ diagonal**: Excellent recovery (corr $= 0.997$, MAE $= 0.009$). Marginal player variances are reliably estimated regardless of loading mixing issues.
- **$\text{var}_b$ vs $\Sigma_b$ diagonal**: Good recovery (corr $= 0.884$, MAE $= 0.030$). Season variance recovery is slightly weaker, as expected with only 12 seasons.

### Variance Decomposition — Real Data (Stage 10)

```{r variance-decomp-tables}
vd <- read_t("11_real_variance_decomposition.csv")

if (!is.null(vd)) {
  top_player <- vd |>
    arrange(desc(prop_a_mean)) |>
    slice_head(n = 8) |>
    select(feature, prop_a_mean, prop_b_mean, prop_e_mean) |>
    mutate(across(starts_with("prop"), \(x) round(x * 100, 1)))
  colnames(top_player) <- c("Feature", "Player %", "Season %", "Residual %")
  show_table(top_player, caption = "Top 8 player-dominated features (sorted by player share)")

  top_season <- vd |>
    arrange(desc(prop_b_mean)) |>
    slice_head(n = 6) |>
    select(feature, prop_a_mean, prop_b_mean, prop_e_mean) |>
    mutate(across(starts_with("prop"), \(x) round(x * 100, 1)))
  colnames(top_season) <- c("Feature", "Player %", "Season %", "Residual %")
  show_table(top_season, caption = "Top 6 season-driven features (sorted by season share)")

  top_resid <- vd |>
    arrange(desc(prop_e_mean)) |>
    slice_head(n = 6) |>
    select(feature, prop_a_mean, prop_b_mean, prop_e_mean) |>
    mutate(across(starts_with("prop"), \(x) round(x * 100, 1)))
  colnames(top_resid) <- c("Feature", "Player %", "Season %", "Residual %")
  show_table(top_resid, caption = "Top 6 residual-dominated features (sorted by residual share)")
}
```

```{r variance-decomp-plot, fig.cap="Variance decomposition by feature and component — low-rank $\\Sigma_a$ model (real data). Features sorted by player share within each group.", out.width="100%", fig.height=14}
p <- fig("11_real_variance_decomposition_lowrank.png")
if (file.exists(p)) knitr::include_graphics(p)
```

### Factor Loading Structure — Real Data

```{r loading-heatmap, fig.cap="$\\Lambda_a$ loading heatmap — real La Liga data (stage 10). Posterior means; rows = features, columns = factors.", out.width="75%", fig.height=14, fig.align="center"}
p <- fig("11_real_lambda_a_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r loading-topn, fig.cap="Top 12 loadings per factor with 90\\% posterior credible intervals — real data.", out.width="100%"}
p <- fig("11_real_lambda_a_top_loadings.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r loading-radar, fig.cap="Mean absolute loading by feature group and factor — real data.", out.width="90%", fig.align="center"}
p <- fig("11_real_lambda_a_group_radar.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r loading-comparison, fig.cap="Factor loading comparison across factors — real data (stage 12).", out.width="100%"}
p <- fig("12_real_loading_comparison.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r group-radar-12, fig.cap="Factor group radar chart — football feature groups (stage 12).", out.width="85%", fig.align="center"}
p <- fig("12_real_factor_group_radar.png")
if (file.exists(p)) knitr::include_graphics(p)
```

**Factor 1 interpretation:** Separates **attacking/finishing players** (high: `touch_in_box` +0.691, `offensive_duels` +0.671, `shots` +0.653, `xg_shot` +0.627, `dribbles` +0.549) from **defensive/possession midfielders** (low: `forward_passes` $-$0.816, `recoveries` $-$0.737, `progressive_passes` $-$0.694, `lateral_passes` $-$0.658, `interceptions` $-$0.615). This is the dominant axis of variation in La Liga player profiles.

**Factor 2 interpretation:** Only `assists` (+0.297) has a credible interval clearly above zero. All other Factor 2 loadings span zero widely. At $Q_a = 2$, Factor 2 is not meaningfully identified in the real data — likely reflects a subtle playmaker/finisher split that requires more data or a stronger anchor.

### Simulation Recovery Scatter (Stage 09)

```{r sim-scatter, fig.cap="$\\Lambda_a$ posterior mean vs simulation truth per factor — stage 09. Dashed line = perfect recovery. Factor 2 recovers almost exactly; Factor 1 does not.", out.width="100%"}
p <- fig("11_sim_lambda_a_vs_truth.png")
if (file.exists(p)) knitr::include_graphics(p)
```

### Chain Diagnostics

#### Stage 09 — Simulated Data

```{r chains-09-diag-trace, fig.cap="Stage 09: $\\lambda_a$ diagonal anchor parameters — trace plots (4 chains).", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_diag_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-09-diag-density, fig.cap="Stage 09: $\\lambda_a$ diagonal anchor parameters — posterior densities by chain.", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_diag_density.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-09-free-trace, fig.cap="Stage 09: top free $\\lambda_a$ parameters (by absolute posterior mean) — trace plots.", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_free_top_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-09-free-density, fig.cap="Stage 09: top free $\\lambda_a$ parameters — posterior densities by chain.", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_free_top_density.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-09-psi, fig.cap="Stage 09: uniqueness parameters $\\psi_a$ (first 6 features) — trace and density.", out.width="100%"}
p_t <- chain_fig("09_sim", "psi_a_trace.png")
p_d <- chain_fig("09_sim", "psi_a_density.png")
if (file.exists(p_t)) knitr::include_graphics(p_t)
if (file.exists(p_d)) knitr::include_graphics(p_d)
```

#### Stage 10 — Real La Liga Data

```{r chains-10-diag-trace, fig.cap="Stage 10 (real data): $\\lambda_a$ diagonal anchors — trace plots.", out.width="100%"}
p <- chain_fig("10_real", "lambda_a_diag_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-10-diag-density, fig.cap="Stage 10 (real data): $\\lambda_a$ diagonal anchors — densities.", out.width="100%"}
p <- chain_fig("10_real", "lambda_a_diag_density.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-10-free-trace, fig.cap="Stage 10 (real data): top free $\\lambda_a$ parameters — trace plots.", out.width="100%"}
p <- chain_fig("10_real", "lambda_a_free_top_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-10-sigma-b, fig.cap="Stage 10 (real data): top season scale parameters $\\sigma_b$ — trace and density.", out.width="100%"}
p_t <- chain_fig("10_real", "sigma_b_top_trace.png")
p_d <- chain_fig("10_real", "sigma_b_top_density.png")
if (file.exists(p_t)) knitr::include_graphics(p_t)
if (file.exists(p_d)) knitr::include_graphics(p_d)
```

### Bugs Fixed This Run

```{r bugs-fixed}
tibble::tribble(
  ~File, ~Bug, ~Fix,
  "11_loading_analysis_and_chains.R",
    "truth\\$psi\\_a\\_sd does not exist in truth object",
    "Replaced with truth\\$Sigma\\_a\\_diag (precomputed diagonal of Sigma\\_a)",
  "11_loading_analysis_and_chains.R",
    "true\\_lambda[row\\_vec, col\\_vec] extracts 106x106 submatrix (11,236 elements) instead of element-wise",
    "Changed to true\\_lambda[cbind(observed\\_index, factor\\_index)]"
) |>
  knitr::kable(booktabs = TRUE, linesep = "",
               caption = "Bugs fixed during this pipeline run",
               escape = FALSE) |>
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"),
                            font_size = 8) |>
  kableExtra::column_spec(1, width = "4cm") |>
  kableExtra::column_spec(2, width = "5.5cm") |>
  kableExtra::column_spec(3, width = "5.5cm")
```

### Notes and Interpretation

**Convergence verdict:** The loading identification problem in Factor 1 persists across both the simulated (stage 09) and real (stage 10) fits. This is consistent with a fundamental sign/scale ambiguity in the first anchor loading under the current constraint. The diagonal variance components and Factor 2 converge cleanly.

**Key football finding:** Factor 1 is the dominant axis of La Liga player variation, cleanly separating attacking/finishing profiles from defensive/possession midfielders. This is stable and interpretable despite the mixing issues, since the posterior median is consistent across chains for the high-magnitude loadings.

**What works well:** Marginal player variances ($\Sigma_a$ diagonal), season variances ($\sigma_b$), residual scales ($\sigma_\varepsilon$), global means ($\mu$), and Factor 2 (both in simulation and partially in real data) all converge reliably.

**What to address next:**
- Investigate stronger anchor constraints or ordering constraints for Factor 1 to resolve the sign/scale ambiguity.
- Consider increasing $Q_a$ or testing $Q_a = 1$ to assess whether a single factor is sufficient for the real data.
- Factor scores for players were not produced (size mismatch in stage 12) — needs a small fix in `12_football_interpretation.R`.
- PSIS-LOO is unreliable for this model (Pareto $k \gg 1$); K-fold cross-validation is the recommended next step for formal model comparison.

---

# Entry Template (Reference for Future Updates)

The following template should be used for each new journal entry appended to this document.
Copy from the `---` separator below and fill in the bracketed fields.

```
---

## Entry [N] — YYYY-MM-DD — [Short descriptive title]

### Configuration

| Parameter | Value |
|---|---|
| Seed | [value] |
| Chains | [value] |
| Iter warmup | [value] |
| Iter sampling | [value] |
| Adapt delta | [value] |
| Max treedepth | [value] |
| Player factor rank Q_a | [value] |

### Stages Run

- **[XX]** — [one-line description of what was run and why]

### Sampler Health

[R code chunk reading from relevant diagnostic CSVs for these stages]

### Recovery Metrics (if sim stage)

[R code chunk reading from 11_sim_recovery_metrics.csv or equivalent]

### Variance Decomposition (if real-data stage)

[Tables + include_graphics for variance decomp plot]

### Loading Structure (if stage 11/12 ran)

[include_graphics for heatmap, top-loadings, radar]

### Chain Diagnostics

[include_graphics for lambda_a_diag and lambda_a_free_top trace/density
 from outputs/figures/11_chain_plots/{model_tag}/]

### Bugs Fixed This Run

[Table of any bugs found and fixed during this run]

### Notes and Interpretation

[Convergence verdict, substantive findings, football interpretation, next steps]
```
