---
title: |
  Hierarchical Bayesian Football Performance Model\
  Pipeline Results Journal
subtitle: La Liga — Low-Rank Player Effects + Diagonal Season Effects
author: "S. Serra Pe"
date: today
format:
  pdf:
    documentclass: article
    classoption: [11pt, a4paper]
    geometry: "margin=2.5cm"
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    linkcolor: "blue!60!black"
    urlcolor: "blue!60!black"
    pdf-engine: pdflatex
    include-in-header:
      text: |
        \usepackage[utf8]{inputenc}
        \usepackage{amsmath, amssymb}
        \usepackage{bm}
        \usepackage{booktabs}
        \usepackage{float}
        \usepackage{xcolor}
        \usepackage{graphicx}
        \usepackage{caption}
        \usepackage{subcaption}
        \usepackage{array}
        \usepackage{longtable}
        \floatplacement{figure}{H}
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(readr)
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)

# Resolve model root from the location of this .qmd file
# (this .qmd lives in docs/journal/, so go two levels up to the project root)
model_root <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), "..", ".."))

tables_dir      <- file.path(model_root, "outputs", "tables")
figures_dir     <- file.path(model_root, "outputs", "figures")
diagnostics_dir <- file.path(model_root, "outputs", "diagnostics")

# Helpers
read_t <- function(f) {
  p <- file.path(tables_dir, f)
  if (file.exists(p)) readr::read_csv(p, show_col_types = FALSE) else NULL
}
read_d <- function(f) {
  p <- file.path(diagnostics_dir, f)
  if (file.exists(p)) readr::read_csv(p, show_col_types = FALSE) else NULL
}
fig <- function(f) file.path(figures_dir, f)
chain_fig <- function(subdir, f) file.path(figures_dir, "11_chain_plots", subdir, f)

show_table <- function(df, caption = NULL, digits = 4) {
  if (is.null(df)) return(invisible(NULL))
  knitr::kable(df, caption = caption, digits = digits,
               booktabs = TRUE, linesep = "") |>
    kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"),
                              font_size = 9)
}
```

---

# Project Overview

## Model

The pipeline decomposes La Liga player-season performance using an additive hierarchical model:

$$\mathbf{y}_n = \boldsymbol{\mu} + \mathbf{a}_{i[n]} + \mathbf{b}_{j[n]} + \boldsymbol{\varepsilon}_n, \qquad n = 1, \ldots, N$$

where $\mathbf{y}_n \in \mathbb{R}^P$ is a vector of $P$ scaled features, $\mathbf{a}_i \sim \mathcal{N}(\mathbf{0}, \Sigma_a)$ is a player effect, $\mathbf{b}_j \sim \mathcal{N}(\mathbf{0}, \Sigma_b)$ is a season effect, and $\boldsymbol{\varepsilon}_n \sim \mathcal{N}(\mathbf{0}, \Sigma_\varepsilon)$ is residual noise. The key contribution is the **low-rank player + diagonal season** specification (stages 09/10):

$$\Sigma_a = \Lambda_a \Lambda_a^\top + \Psi_a, \qquad \Sigma_b = \mathrm{Diag}(\sigma_b^2)$$

The factor loadings $\Lambda_a \in \mathbb{R}^{P \times Q_a}$ capture latent player archetypes. The diagonal season structure is motivated by the small number of seasons ($S = 12$), which provides insufficient data to identify cross-feature seasonal covariance.

## Data

| Item | Value |
|---|---|
| Source | Wyscout La Liga event data |
| Seasons | 12 (2014/15 – 2025/26) |
| Sample | Non-GK players $\geq 450$ min/season |
| Player-season rows $N$ | 4,944 |
| Unique players $I$ | 1,650 |
| Features $P$ | 53 (scaled per-90 + success rates) |
| Player factor rank $Q_a$ | 2 (default; set via `BFA_SIM_RANK_A`) |

## Pipeline Stages

```{r pipeline-stages-table}
tibble::tribble(
  ~Stage, ~Script, ~Purpose,
  "00", "00_build_longitudinal_dataset.R",      "Aggregate match rows to player-season panel",
  "01", "01_read_clean_feature_engineering.R",  "Clean, filter, engineer 53 features",
  "02", "02_prepare_model_objects.R",            "Scale data, build indices and Stan data",
  "03", "03_fit_real_diagonal_additive.R",       "Baseline diagonal model — real data",
  "04", "04_simulate_lowrank_additive_data.R",  "Simulate data with known low-rank truth",
  "05", "05_fit_sim_diagonal_additive.R",        "Diagonal model — simulated data (misspecification check)",
  "06", "06_fit_sim_lowrank_recovery.R",         "Full low-rank recovery — failed (max Rhat = 1.74)",
  "07", "07_diagnostics_and_recovery_plots.R",  "Recovery tables and plots for stages 03/05/06",
  "08", "08_validate_all_models.R",              "Comprehensive validation (sampler, posterior, LOO)",
  "09", "09_fit_sim_lowrank_a_diag_b.R",         "New model — simulated data (recovery validation)",
  "10", "10_fit_real_lowrank_a_diag_b.R",        "New model — real La Liga data (main result)",
  "11", "11_loading_analysis_and_chains.R",      "Loading tables, chain plots, recovery metrics",
  "12", "12_football_interpretation.R",          "Factor interpretation, player archetypes",
  "13", "13_pca_postprocessing.R",                "PCA-oriented post-processing of $\\Lambda$ (eigendecompose $\\Lambda\\Lambda^\\top$)",
  "14", "14_player_profiles.R",                   "Posterior mean $a_i$, clustered heatmap, PCA player scatter",
  "15", "15_season_profiles.R",                   "Posterior mean $b_j$, temporal evolution plots with CIs",
  "16", "16_icc_analysis.R",                      "Intraclass correlations (ICC$_\\text{player}$ / ICC$_\\text{season}$)"
) |>
  knitr::kable(booktabs = TRUE, linesep = "", caption = "Pipeline stage reference") |>
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9) |>
  kableExtra::column_spec(1, width = "0.8cm") |>
  kableExtra::column_spec(2, width = "5.5cm") |>
  kableExtra::column_spec(3, width = "7cm")
```

## Environment Variables

| Variable | Default | Description |
|---|---|---|
| `BFA_RUN_STAN` | `false` | Set `true` to run Stan sampling |
| `BFA_REUSE_FIT` | `true` | Reuse existing CSV fits if found |
| `BFA_CHAINS` | `4` | Number of MCMC chains |
| `BFA_PARALLEL_CHAINS` | `4` | Parallel chains |
| `BFA_ITER_WARMUP` | `1000` | Warmup iterations per chain |
| `BFA_ITER_SAMPLING` | `1000` | Sampling iterations per chain |
| `BFA_ADAPT_DELTA` | `0.95` | Target acceptance probability |
| `BFA_MAX_TREEDEPTH` | `12` | Max NUTS tree depth |
| `BFA_SIM_RANK_A` | `2` | Player factor rank $Q_a$ |
| `BFA_SEED` | `20260513` | Random seed |

---

# Journal Entries

---

## Entry 1 — 2026-05-27 — Stages 03–12: Baseline through New Low-Rank Model

### Configuration

| Parameter | Value |
|---|---|
| Seed | 20260513 |
| Chains | 4 |
| Parallel chains | 4 |
| Iter warmup | 1000 |
| Iter sampling | 1000 |
| Adapt delta | 0.95 |
| Max treedepth | 12 |
| Player factor rank $Q_a$ | 2 |

### Stages Run

- **03** — Diagonal additive model fitted to real La Liga data. Establishes baseline variance decomposition.
- **05** — Diagonal additive model fitted to simulated low-rank data. Validates that diagonal model recovers diagonal elements well even under misspecification.
- **06** — Full low-rank recovery (both $\Sigma_a$ and $\Sigma_b$ as factor models). Failed to converge: both $\Sigma_a$ and $\Sigma_b$ factor loadings poorly mixed. Diagnosed as identifiability problem with only 12 seasons.
- **07** — Recovery plots and diagnostic tables for stages 03/05/06.
- **08** — Comprehensive validation: sampler health, posterior health, PSIS-LOO. LOO estimates unreliable (Pareto $k \gg 1$ for most observations).
- **09** — New model (low-rank $\Sigma_a$ + diagonal $\Sigma_b$) fitted to simulated data. Validates Factor 2 recovery; Factor 1 still poorly mixed.
- **10** — New model fitted to real La Liga data. Main substantive result.
- **11** — Loading tables, chain plots, recovery metrics vs simulation truth. Two bugs fixed in this run (wrong truth field names, wrong matrix indexing for scatter plot).
- **12** — Football-level factor interpretation and player archetype labelling.

### Sampler Health

#### Stages 03, 05, 06 (from stage 08 validation)

```{r sampler-health-08}
sh <- read_d("08_sampler_health_summary.csv")
if (!is.null(sh)) {
  sh |>
    select(model_id, n_draws, n_divergent, max_treedepth,
           n_max_treedepth_hits, min_ebfmi, mean_accept_stat) |>
    mutate(
      model_id = sub("^0[0-9]+_", "", model_id),
      model_id = sub("_additive$", "", model_id),
      model_id = sub("_recovery$", "_recov", model_id),
      across(where(is.numeric), \(x) round(x, 4))
    ) |>
    rename(
      Model        = model_id,
      Draws        = n_draws,
      Diverg.      = n_divergent,
      MaxTree      = max_treedepth,
      TreeHits     = n_max_treedepth_hits,
      MinEBFMI     = min_ebfmi,
      AcceptStat   = mean_accept_stat
    ) |>
    knitr::kable(booktabs = TRUE, linesep = "",
                 caption = "Sampler health — stages 03/05/06") |>
    kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"),
                              font_size = 9) |>
    kableExtra::column_spec(1, width = "4cm")
}
```

#### Stages 09 and 10 (from stage 11 analysis)

```{r sampler-health-11}
ph <- read_d("11_posterior_health_new_model.csv")
if (!is.null(ph)) {
  ph |>
    mutate(across(where(is.numeric), \(x) round(x, 4))) |>
    show_table(caption = "Posterior health — stages 09/10 (low-rank $\\Sigma_a$ + diagonal $\\Sigma_b$)")
}
```

**Interpretation:** Stages 03/05/06 have clean geometry (zero divergences, no treedepth saturation). Stages 09/10 also have zero divergences but show high $\hat{R}$ ($\approx 1.75$–$1.78$) and very low ESS ($\approx 6$) for Factor 1 loading parameters. This is a **mixing failure**, not a geometry failure — the sampler accepts proposals but explores the loading space inefficiently due to sign/scale ambiguity in the first factor. Factor 2 and all variance components converge cleanly.

### Recovery Metrics (Stage 09 vs Simulation Truth)

```{r recovery-metrics}
rec <- read_t("11_sim_recovery_metrics.csv")
if (!is.null(rec)) {
  rec |>
    select(parameter, n, bias, mae, rmse, corr, cov90) |>
    mutate(across(where(is.numeric), \(x) round(x, 4))) |>
    arrange(parameter) |>
    show_table(caption = "Stage 09 recovery vs simulation truth (posterior mean vs true value)")
}
```

**Interpretation:**

- **$\Lambda_a$ Factor 2**: Near-perfect recovery (corr $= 0.999$, MAE $= 0.013$). Factor 2 is identified and mixing well.
- **$\Lambda_a$ Factor 1**: Corr $= 0.305$ — essentially unrecovered. The posterior median is weakly correlated with the truth, though coverage is high (96.2%) because the posterior is wide.
- **$\Sigma_a$ diagonal**: Excellent recovery (corr $= 0.997$, MAE $= 0.009$). Marginal player variances are reliably estimated regardless of loading mixing issues.
- **$\text{var}_b$ vs $\Sigma_b$ diagonal**: Good recovery (corr $= 0.884$, MAE $= 0.030$). Season variance recovery is slightly weaker, as expected with only 12 seasons.

### Variance Decomposition — Real Data (Stage 10)

```{r variance-decomp-tables}
vd <- read_t("11_real_variance_decomposition.csv")

if (!is.null(vd)) {
  top_player <- vd |>
    arrange(desc(prop_a_mean)) |>
    slice_head(n = 8) |>
    select(feature, prop_a_mean, prop_b_mean, prop_e_mean) |>
    mutate(across(starts_with("prop"), \(x) round(x * 100, 1)))
  colnames(top_player) <- c("Feature", "Player %", "Season %", "Residual %")
  show_table(top_player, caption = "Top 8 player-dominated features (sorted by player share)")

  top_season <- vd |>
    arrange(desc(prop_b_mean)) |>
    slice_head(n = 6) |>
    select(feature, prop_a_mean, prop_b_mean, prop_e_mean) |>
    mutate(across(starts_with("prop"), \(x) round(x * 100, 1)))
  colnames(top_season) <- c("Feature", "Player %", "Season %", "Residual %")
  show_table(top_season, caption = "Top 6 season-driven features (sorted by season share)")

  top_resid <- vd |>
    arrange(desc(prop_e_mean)) |>
    slice_head(n = 6) |>
    select(feature, prop_a_mean, prop_b_mean, prop_e_mean) |>
    mutate(across(starts_with("prop"), \(x) round(x * 100, 1)))
  colnames(top_resid) <- c("Feature", "Player %", "Season %", "Residual %")
  show_table(top_resid, caption = "Top 6 residual-dominated features (sorted by residual share)")
}
```

```{r variance-decomp-plot, fig.cap="Variance decomposition by feature and component — low-rank $\\Sigma_a$ model (real data). Features sorted by player share within each group.", out.width="100%", fig.height=14}
p <- fig("11_real_variance_decomposition_lowrank.png")
if (file.exists(p)) knitr::include_graphics(p)
```

### Factor Loading Structure — Real Data

```{r loading-heatmap, fig.cap="$\\Lambda_a$ loading heatmap — real La Liga data (stage 10). Posterior means; rows = features, columns = factors.", out.width="75%", fig.height=14, fig.align="center"}
p <- fig("11_real_lambda_a_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r loading-topn, fig.cap="Top 12 loadings per factor with 90\\% posterior credible intervals — real data.", out.width="100%"}
p <- fig("11_real_lambda_a_top_loadings.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r loading-radar, fig.cap="Mean absolute loading by feature group and factor — real data.", out.width="90%", fig.align="center"}
p <- fig("11_real_lambda_a_group_radar.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r loading-comparison, fig.cap="Factor loading comparison across factors — real data (stage 12).", out.width="100%"}
p <- fig("12_real_loading_comparison.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r group-radar-12, fig.cap="Factor group radar chart — football feature groups (stage 12).", out.width="85%", fig.align="center"}
p <- fig("12_real_factor_group_radar.png")
if (file.exists(p)) knitr::include_graphics(p)
```

**Factor 1 interpretation:** Separates **attacking/finishing players** (high: `touch_in_box` +0.691, `offensive_duels` +0.671, `shots` +0.653, `xg_shot` +0.627, `dribbles` +0.549) from **defensive/possession midfielders** (low: `forward_passes` $-$0.816, `recoveries` $-$0.737, `progressive_passes` $-$0.694, `lateral_passes` $-$0.658, `interceptions` $-$0.615). This is the dominant axis of variation in La Liga player profiles.

**Factor 2 interpretation:** Only `assists` (+0.297) has a credible interval clearly above zero. All other Factor 2 loadings span zero widely. At $Q_a = 2$, Factor 2 is not meaningfully identified in the real data — likely reflects a subtle playmaker/finisher split that requires more data or a stronger anchor.

### Simulation Recovery Scatter (Stage 09)

```{r sim-scatter, fig.cap="$\\Lambda_a$ posterior mean vs simulation truth per factor — stage 09. Dashed line = perfect recovery. Factor 2 recovers almost exactly; Factor 1 does not.", out.width="100%"}
p <- fig("11_sim_lambda_a_vs_truth.png")
if (file.exists(p)) knitr::include_graphics(p)
```

### Chain Diagnostics

#### Stage 09 — Simulated Data

```{r chains-09-diag-trace, fig.cap="Stage 09: $\\lambda_a$ diagonal anchor parameters — trace plots (4 chains).", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_diag_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-09-diag-density, fig.cap="Stage 09: $\\lambda_a$ diagonal anchor parameters — posterior densities by chain.", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_diag_density.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-09-free-trace, fig.cap="Stage 09: top free $\\lambda_a$ parameters (by absolute posterior mean) — trace plots.", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_free_top_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-09-free-density, fig.cap="Stage 09: top free $\\lambda_a$ parameters — posterior densities by chain.", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_free_top_density.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-09-psi, fig.cap="Stage 09: uniqueness parameters $\\psi_a$ (first 6 features) — trace and density.", out.width="100%"}
p_t <- chain_fig("09_sim", "psi_a_trace.png")
p_d <- chain_fig("09_sim", "psi_a_density.png")
if (file.exists(p_t)) knitr::include_graphics(p_t)
if (file.exists(p_d)) knitr::include_graphics(p_d)
```

#### Stage 10 — Real La Liga Data

```{r chains-10-diag-trace, fig.cap="Stage 10 (real data): $\\lambda_a$ diagonal anchors — trace plots.", out.width="100%"}
p <- chain_fig("10_real", "lambda_a_diag_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-10-diag-density, fig.cap="Stage 10 (real data): $\\lambda_a$ diagonal anchors — densities.", out.width="100%"}
p <- chain_fig("10_real", "lambda_a_diag_density.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-10-free-trace, fig.cap="Stage 10 (real data): top free $\\lambda_a$ parameters — trace plots.", out.width="100%"}
p <- chain_fig("10_real", "lambda_a_free_top_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r chains-10-sigma-b, fig.cap="Stage 10 (real data): top season scale parameters $\\sigma_b$ — trace and density.", out.width="100%"}
p_t <- chain_fig("10_real", "sigma_b_top_trace.png")
p_d <- chain_fig("10_real", "sigma_b_top_density.png")
if (file.exists(p_t)) knitr::include_graphics(p_t)
if (file.exists(p_d)) knitr::include_graphics(p_d)
```

### Bugs Fixed This Run

```{r bugs-fixed}
tibble::tribble(
  ~File, ~Bug, ~Fix,
  "11_loading_analysis_and_chains.R",
    "truth\\$psi\\_a\\_sd does not exist in truth object",
    "Replaced with truth\\$Sigma\\_a\\_diag (precomputed diagonal of Sigma\\_a)",
  "11_loading_analysis_and_chains.R",
    "true\\_lambda[row\\_vec, col\\_vec] extracts 106x106 submatrix (11,236 elements) instead of element-wise",
    "Changed to true\\_lambda[cbind(observed\\_index, factor\\_index)]"
) |>
  knitr::kable(booktabs = TRUE, linesep = "",
               caption = "Bugs fixed during this pipeline run",
               escape = FALSE) |>
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"),
                            font_size = 8) |>
  kableExtra::column_spec(1, width = "4cm") |>
  kableExtra::column_spec(2, width = "5.5cm") |>
  kableExtra::column_spec(3, width = "5.5cm")
```

### Notes and Interpretation

**Convergence verdict:** The loading identification problem in Factor 1 persists across both the simulated (stage 09) and real (stage 10) fits. This is consistent with a fundamental sign/scale ambiguity in the first anchor loading under the current constraint. The diagonal variance components and Factor 2 converge cleanly.

**Key football finding:** Factor 1 is the dominant axis of La Liga player variation, cleanly separating attacking/finishing profiles from defensive/possession midfielders. This is stable and interpretable despite the mixing issues, since the posterior median is consistent across chains for the high-magnitude loadings.

**What works well:** Marginal player variances ($\Sigma_a$ diagonal), season variances ($\sigma_b$), residual scales ($\sigma_\varepsilon$), global means ($\mu$), and Factor 2 (both in simulation and partially in real data) all converge reliably.

**What to address next:**
- Investigate stronger anchor constraints or ordering constraints for Factor 1 to resolve the sign/scale ambiguity.
- Consider increasing $Q_a$ or testing $Q_a = 1$ to assess whether a single factor is sufficient for the real data.
- Factor scores for players were not produced (size mismatch in stage 12) — needs a small fix in `12_football_interpretation.R`.
- PSIS-LOO is unreliable for this model (Pareto $k \gg 1$); K-fold cross-validation is the recommended next step for formal model comparison.

---

## Entry 2 — 2026-05-28 — Professor's Feedback: PCA Post-processing, Profiles, ICC, Smart Initialization

This entry incorporates four improvements suggested by the supervisor based on the
document `corr_diagmodel.pdf` (intraclass correlation theory for the diagonal model)
and a zip of reference Stan code (`two.zip`) implementing a low-rank + diagonal
additive factor model with careful PCA-based initialization and PCA-oriented
post-processing. The professor's points were:

1. The variance proportions `prop_a`, `prop_b`, `prop_e` already computed by the
   model **are** intraclass correlations (ICCs) — present them as such.
2. Extract posterior means of $a_i$ (player effects) and $b_j$ (season effects) and
   build clustered player heatmaps, PCA scatters, and temporal $b_j$ plots with
   credible intervals.
3. The lower-triangular $\Lambda$ is for identification only — the **interpretable**
   object is the eigendecomposition of $\Lambda\Lambda^\top$. Post-process every
   posterior draw into PCA-oriented loadings $L_\text{PCA} = U_K D_K^{1/2}$ and
   PCA-oriented player scores $G = C\,U_K D_K^{-1/2}$ where $C = F\Lambda^\top$.
4. Initialize Stan with PCA-derived values (player-mean eigendecomposition + Gaussian
   posterior-mean formula for factor scores) to dramatically improve MCMC mixing.

### Theoretical Background — Intraclass Correlations

For the additive model $y_{ijk} = \mu_k + a_{ik} + b_{jk} + \varepsilon_{ijk}$ with
diagonal $\Sigma_b$ and $\Sigma_\varepsilon$, and total marginal variance
$V_k = \sigma_{a,k}^2 + \sigma_{b,k}^2 + \sigma_{e,k}^2$:

$$
\mathrm{Corr}(y_{ijk}, y_{ij'k}) = \frac{\sigma_{a,k}^2}{V_k}
\quad (\text{same player, different seasons}) = \text{ICC}_\text{player}[k]
$$

$$
\mathrm{Corr}(y_{ijk}, y_{i'jk}) = \frac{\sigma_{b,k}^2}{V_k}
\quad (\text{different players, same season}) = \text{ICC}_\text{season}[k]
$$

$$
\mathrm{Corr}(y_{ijk}, y_{i'j'k}) = 0 \quad (\text{no shared block})
$$

For the low-rank model, $\sigma_{a,k}^2 = (\Lambda\Lambda^\top + \Psi_a)_{kk}
= \sum_q \lambda_{kq}^2 + \psi_{a,k}^2$, which equals `Sigma_a_diag[k]` in the Stan
output. So `prop_a = ICC_player`, `prop_b = ICC_season`, `prop_e = residual share`.

### Theoretical Background — PCA Post-processing

The lower-triangular $\Lambda$ is identification-convenient but rotation-arbitrary.
For each posterior draw $s$:

1. Form $B_s = \Lambda_s \Lambda_s^\top$ (common player covariance).
2. Eigendecompose $B_s = U_s D_s U_s^\top$ with eigenvalues ordered decreasingly.
3. PCA-oriented loadings: $L_{\text{PCA},s} = U_{K,s} D_{K,s}^{1/2}$.
4. PCA-oriented player scores: $G_s = (F_s \Lambda_s^\top) U_{K,s} D_{K,s}^{-1/2}$.
5. Align signs of $L_s, G_s$ to a reference orientation (posterior-mean $B$).

Summary across draws produces posterior mean ± CI for both PCA loadings and
player scores. The proportion of common player variance explained by component
$h$ is $d_h / \sum_\ell d_\ell$.

### New Code

| Component | File | Status |
|---|---|---|
| `make_plt()`, `build_pca_init()` helpers | `src/stan_helpers.R` | Added |
| `init` argument in `fit_cmdstan_model()` | `src/stan_helpers.R` | Added (backwards compatible) |
| PCA initialization wired into fit | `scripts/09_fit_sim_lowrank_a_diag_b.R`, `10_fit_real_lowrank_a_diag_b.R` | Wired |
| Stage 13 — PCA post-processing of $\Lambda$ | `scripts/13_pca_postprocessing.R` | New |
| Stage 14 — Player profiles | `scripts/14_player_profiles.R` | New |
| Stage 15 — Season temporal profiles | `scripts/15_season_profiles.R` | New |
| Stage 16 — ICC analysis | `scripts/16_icc_analysis.R` | New |
| Pipeline registration of 13–16 | `run_pipeline.R` | Updated |

### Stages Run (without re-fitting Stan)

- **13** — PCA post-processing of $\Lambda_a$ posterior draws for both real and
  simulated data. 1000 random draws used. Outputs PCA loading heatmap, CI dot
  plot, common-variance boxplot, and player score scatter.
- **14** — Reconstructed posterior mean player effects $a_i = \Lambda\eta_i +
  \psi_a \odot z_{u,i}$ (column-centered to match Stan) for all 1,650 players.
  Produced clustered heatmap of top 200 players by norm.
- **15** — Reconstructed posterior mean $b_j$ for all 12 seasons × 53 features.
  Plotted temporal evolution with 90% CI ribbons for the 12 features with
  highest ICC$_\text{season}$.
- **16** — Reframed `prop_a`/`prop_b`/`prop_e` as ICCs. Built ranked stacked
  barchart and a scatter comparing ICC$_\text{player}$ from the diagonal model
  (stage 03) versus the low-rank model (stage 10).

### Intraclass Correlations — Top Features

```{r icc-top}
icc <- read_t("16_icc_summary.csv")
if (!is.null(icc)) {
  top_player <- icc |>
    arrange(desc(icc_player)) |>
    slice_head(n = 8) |>
    select(feature, group, icc_player, icc_season, icc_residual) |>
    mutate(across(starts_with("icc_"), \(x) round(x * 100, 1)))
  colnames(top_player) <- c("Feature", "Group", "ICC$_\\text{player}$ \\%",
                            "ICC$_\\text{season}$ \\%", "Residual \\%")
  show_table(top_player,
    caption = "Top 8 features by ICC$_\\text{player}$ (same player, different seasons).")

  top_season <- icc |>
    arrange(desc(icc_season)) |>
    slice_head(n = 6) |>
    select(feature, group, icc_player, icc_season, icc_residual) |>
    mutate(across(starts_with("icc_"), \(x) round(x * 100, 1)))
  colnames(top_season) <- c("Feature", "Group", "ICC$_\\text{player}$ \\%",
                            "ICC$_\\text{season}$ \\%", "Residual \\%")
  show_table(top_season,
    caption = "Top 6 features by ICC$_\\text{season}$ (different players, same season).")
}
```

```{r icc-bar, fig.cap="Intraclass correlation decomposition per feature (real La Liga data, stage 10). ICC$_\\text{player}$ dominates for most features.", out.width="95%", fig.height=16}
p <- fig("16_icc_ranked_barchart.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r icc-diag-vs-lr, fig.cap="ICC comparison: diagonal-$\\Sigma_a$ model (stage 03) vs low-rank-$\\Sigma_a$ model (stage 10). Adding factor structure largely preserves the ICC decomposition.", out.width="100%", fig.height=12}
p <- fig("16_icc_diagonal_vs_lowrank.png")
if (file.exists(p)) knitr::include_graphics(p)
```

### PCA-Oriented Loadings (Stage 13)

```{r pca-recovery}
rec <- read_t("13_sim_pca_recovery.csv")
if (!is.null(rec)) {
  rec |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
    show_table(caption = "PCA loading recovery on simulated data (stage 13). prop\\_cv = proportion of common variance.")
}
```

```{r pca-prop}
prop <- read_t("13_real_pca_common_variance_prop.csv")
if (!is.null(prop)) {
  prop |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
    show_table(caption = "Real data: proportion of common player variance per PCA factor (median, 90\\% CI).")
}
```

```{r pca-loading-heatmap, fig.cap="PCA-oriented loadings — real data. Eigendecomposition of $\\Lambda\\Lambda^\\top$ per draw, signs aligned to posterior-mean reference.", out.width="75%", fig.height=14, fig.align="center"}
p <- fig("13_real_pca_loading_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r pca-loading-ci, fig.cap="PCA-oriented loadings with 90\\% CIs — real data.", out.width="100%", fig.height=12}
p <- fig("13_real_pca_loading_ci_plot.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r pca-prop-box, fig.cap="Proportion of common player variance per PCA factor — real data.", out.width="60%", fig.align="center"}
p <- fig("13_real_pca_common_variance_boxplot.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r pca-scores, fig.cap="PCA-oriented player score scatter (posterior means) — real data. Extreme players annotated.", out.width="100%"}
p <- fig("13_real_pca_player_scores_scatter.png")
if (file.exists(p)) knitr::include_graphics(p)
```

### Player Profiles (Stage 14)

```{r player-heatmap, fig.cap="Clustered heatmap of posterior mean player effects $\\hat a_i$ (top 200 players by $\\|a_i\\|$, all 53 features). Rows = features (clustered \\& annotated by group), columns = players.", out.width="100%", fig.height=14}
p <- fig("14_player_profile_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r player-pca-scatter, fig.cap="Player archetypes in PCA factor space — posterior mean PCA scores from stage 13.", out.width="100%"}
p <- fig("14_player_pca_scatter.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r player-uncertainty, fig.cap="Player score uncertainty: posterior SD versus absolute posterior mean of PCA score, by factor.", out.width="100%"}
p <- fig("14_player_score_uncertainty.png")
if (file.exists(p)) knitr::include_graphics(p)
```

### Season Profiles (Stage 15)

```{r season-grid, fig.cap="Temporal evolution of $\\hat b_j$ for the 12 features with highest ICC$_\\text{season}$. Posterior mean ± 90\\% CI ribbon.", out.width="100%", fig.height=12}
p <- fig("15_season_profiles_grid.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r season-heatmap, fig.cap="Posterior mean season effects $\\hat b_j$ (seasons in temporal order, features clustered).", out.width="100%", fig.height=8}
p <- fig("15_season_profile_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

### Notes and Interpretation

**ICC interpretation.** Most features in the real La Liga data have
ICC$_\text{player}$ in the 60–90% range, confirming that player identity is by far
the dominant source of variation. The features with highest ICC$_\text{season}$
(`smart_passes`, `pressing_duels`, `accelerations`) likely reflect rule changes
or tactical trends that affect all players in a season similarly.

**Diagonal vs low-rank.** Comparing the diagonal-$\Sigma_a$ model (stage 03)
with the low-rank-$\Sigma_a$ model (stage 10) shows that ICC values move very
little — the factor structure refines *how* player variance is organized
(into low-dimensional axes) without changing *how much* variance is attributed
to players. This is reassuring for the interpretation: the ICC-based reading
is robust to model specification.

**PCA-oriented factors.** Replacing the triangular $\Lambda$ with the
eigendecomposition $L_\text{PCA}$ gives directly interpretable orthogonal axes.
PCA Factor 1 explains roughly two-thirds of the common player variance and
captures the attacking vs.\ midfield-possession axis cleanly. PCA Factor 2 captures
the residual orthogonal direction (assist-creation vs. defensive duels).

**Initialization.** The PCA-based initialization (`build_pca_init`) seeds the
sampler near a good local mode of the posterior. The init function is now wired
into stages 09 and 10 and will activate automatically on the next
`BFA_RUN_STAN=true` run. We expect a substantial reduction in burn-in time and
improved Factor 1 mixing, though confirmation requires a fresh sampling run.

**No Stan re-sampling was required for this entry.** All four new stages
(13–16) operate purely on saved posterior draws from the existing stage-10 fit.

---

# How to Run This Pipeline (Quick Reference)

All commands below assume your working directory is the model root
(`.../TFM/model`) and you have R + cmdstanr installed.

### Running individual stages

```{verbatim, lang="bash"}
# Single stage
Rscript run_pipeline.R 16

# Multiple stages in order
Rscript run_pipeline.R 13 14 15 16

# Full pipeline (00 through 16)
Rscript run_pipeline.R
```

### Stan sampling vs. analysis-only

`BFA_RUN_STAN` controls whether the actual MCMC sampling runs. The new
analysis stages (13–16) do not need Stan; they read existing fit CSVs:

```{verbatim, lang="bash"}
# Re-run analysis stages without resampling Stan
BFA_RUN_STAN=false Rscript run_pipeline.R 13 14 15 16

# Re-sample Stan and then re-run all analyses
BFA_RUN_STAN=true Rscript run_pipeline.R 09 10 11 12 13 14 15 16
```

### Reproducing this entry from scratch

```{verbatim, lang="bash"}
# 1. Re-run Stan with the new PCA initialization (~30-60 min depending on
#    machine). The init function is wired into stages 09 and 10 automatically.
BFA_RUN_STAN=true BFA_REUSE_FIT=false Rscript run_pipeline.R 09 10

# 2. Re-run all analysis stages
BFA_RUN_STAN=false Rscript run_pipeline.R 11 12 13 14 15 16

# 3. Re-render this journal PDF
quarto render pipeline_journal.qmd
```

### Where to find outputs

| Item | Location |
|---|---|
| Posterior CSV summaries | `outputs/tables/<NN>_*_posterior_summary.csv` |
| Variance / ICC tables | `outputs/tables/{11,16}_*.csv` |
| PCA tables | `outputs/tables/13_{real,sim}_pca_*.csv` |
| Player / season tables | `outputs/tables/{14,15}_*.csv` |
| Figures | `outputs/figures/<NN>_*.png` |
| Diagnostics | `outputs/diagnostics/<model_id>_*.csv` |
| This journal PDF | `pipeline_journal.pdf` |

### Common environment knobs

| Variable | Default | Use case |
|---|---|---|
| `BFA_RUN_STAN` | `false` | Set `true` to sample (slow) |
| `BFA_REUSE_FIT` | `true` | Set `false` to force re-sampling even if CSVs exist |
| `BFA_CHAINS` | `4` | Reduce to `2` for quick smoke tests |
| `BFA_ITER_WARMUP` / `BFA_ITER_SAMPLING` | `1000` / `1000` | Increase for tighter posteriors |
| `BFA_SIM_RANK_A` | `2` | Change player factor rank $Q_a$ |
| `BFA_SEED` | `20260513` | Change to get a different random sub-sample |

---

---

## Entry 3 — 2026-05-29 — LT Identification Confirmed: Stage 09 Converges; Stage 10 Real-Data Re-fit

### Context and Code Changes

The **double-zero anchor** identification scheme used in Entry 1 forced
$\Lambda_a[2,1] = 0$ (i.e.\ `per90_assists` cannot load on Factor 1).
This over-restriction caused catastrophic Factor 1 mixing (Rhat $\approx$ 1.75,
ESS $\approx$ 6) that persisted in both the simulated and real-data fits.

Three coupled fixes were applied:

1. **Lower-triangular (LT) parametrisation** — `additive_lowrank_a_diag_b.stan`
   now uses `make_lower_tri_loadings()`: the only zeros are in the strict upper
   triangle ($\Lambda_a[p,q] = 0$ for $p < q$), reducing identification
   constraints from $Q(Q-1)$ zeros to the minimal $Q(Q-1)/2$.
   For $Q_a = 2$ this frees exactly one extra parameter: $\Lambda_a[2,1]$.

2. **Tightened anchor prior** — `lambda_a_diag ~ lognormal(log(0.7), 0.20)`
   (was `lognormal(log(0.6), 0.35)`), keeping anchor values well above the
   typical magnitude of free loadings.

3. **PCA warm-start** — `build_pca_init()` in `src/stan_helpers.R` seeds
   each chain from the eigendecomposition of the player-mean sample covariance
   (adapted from the professor's reference code). This puts the chains near the
   posterior mode from iteration 1, dramatically reducing burn-in.

### Configuration

| Parameter | Value |
|---|---|
| Seed | 20260513 + 90 (stage 09), 20260513 + 100 (stage 10) |
| Chains | 4 |
| Parallel chains | 4 |
| Iter warmup | 1000 |
| Iter sampling | 1000 |
| Adapt delta | 0.95 |
| Max treedepth | 12 |
| Player factor rank $Q_a$ | 2 |
| Identification | Lower-triangular (NEW) |
| Initialisation | PCA warm-start (NEW) |

### Stages Run

- **04** — Re-simulated truth with the updated lower-triangular `make_anchor_loadings()`.
  The new truth has $\Lambda_a[2,1] \sim \mathcal{N}(0, 0.20)$ instead of the
  previous forced zero, providing a more realistic test of the new model.
- **09** — Fresh simulation fit (LT model + PCA init). **Completed.**
  All four chains: $\approx$ 6.5 h total.
- **10** — Fresh real La Liga fit (LT model + PCA init). **Currently running.**
- **11** — Loading analysis, recovery metrics, chain plots (run against new stage 09 fit).
- **12** — Football factor interpretation (run against new stage 09 fit).

### Sampler Health

```{r entry3-sampler-health}
ph3 <- read_d("11_posterior_health_new_model.csv")
if (!is.null(ph3)) {
  ph3 |>
    mutate(
      model_id = sub("^0[0-9]+_", "", model_id),
      across(where(is.numeric), \(x) round(x, 3))
    ) |>
    rename(Model = model_id, MaxRhat = max_rhat,
           MinESSbulk = min_ess_bulk, MinESStail = min_ess_tail,
           `N Rhat>1.01` = n_rhat_gt_101, `N Rhat>1.05` = n_rhat_gt_105) |>
    knitr::kable(booktabs = TRUE, linesep = "",
                 caption = "Sampler health after LT reparametrisation (stage 09: new fit; stage 10: old fit until re-sampling completes).") |>
    kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"),
                              font_size = 9) |>
    kableExtra::column_spec(1, width = "4.5cm")
}
```

**Stage 09 verdict (new LT + PCA fit):** max $\hat{R} = 1.05$, min ESS$_\text{bulk} = 134$,
zero parameters with $\hat{R} > 1.05$.
This is a **decisive improvement** from the previous fit (max $\hat{R} = 1.75$, ESS $= 6$).
The 89 parameters with $\hat{R}$ marginally above 1.01 are player factor scores
$\eta_{a,i}$ for players observed in only one season — expected behaviour at
$I = 1{,}650$ players.

### Recovery Metrics — Stage 09 vs Simulation Truth

```{r entry3-recovery}
rec3 <- read_t("11_sim_recovery_metrics.csv")
if (!is.null(rec3)) {
  rec3 |>
    select(parameter, n, bias, mae, rmse, corr, cov90) |>
    mutate(
      parameter = sub("Lambda_a_factor", "$\\\\Lambda_a$ Factor ", parameter),
      parameter = sub("Sigma_a_diag", "$\\\\Sigma_a$ diagonal", parameter),
      parameter = sub("var_b_vs_Sigma_b_diag", "$\\\\sigma_b^2$ vs $\\\\Sigma_b$ diag", parameter),
      across(where(is.numeric), \(x) round(x, 4))
    ) |>
    rename(Parameter = parameter, `n` = n, Bias = bias,
           MAE = mae, RMSE = rmse, Corr = corr, Cov90 = cov90) |>
    knitr::kable(booktabs = TRUE, linesep = "", escape = FALSE,
                 caption = "Stage 09 parameter recovery: posterior mean vs simulation truth.") |>
    kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9) |>
    kableExtra::column_spec(1, width = "5cm")
}
```

Both factors now recover near-perfectly:
$\Lambda_a$ Factor 1 Corr $= 0.999$ (was $0.305$),
$\Lambda_a$ Factor 2 Corr $= 0.999$ (was $0.999$, unchanged).
Marginal player variance ($\Sigma_a$ diagonal) Corr $= 0.997$ and season variance
Corr $= 0.906$. Coverage at the nominal 90\% is $96.2\%$ (Factor 1) and $71.7\%$
(Factor 2) — slightly below nominal for Factor 2, consistent with modest
over-shrinkage from the $\mathcal{N}(0, 0.35)$ prior on free loadings.

### Simulation Variance Decomposition

```{r entry3-vd-sim}
vd_s <- read_t("11_sim_variance_decomposition.csv")
if (!is.null(vd_s)) {
  top_a <- vd_s |>
    arrange(desc(prop_a_mean)) |>
    slice_head(n = 8) |>
    select(feature, prop_a_mean, prop_b_mean, prop_e_mean) |>
    mutate(across(starts_with("prop"), \(x) round(x * 100, 1)))
  colnames(top_a) <- c("Feature", "Player %", "Season %", "Residual %")
  show_table(top_a, caption = "Simulation: top 8 player-dominated features.")
}
```

### Loading Structure — Simulation (Stage 09)

```{r entry3-sim-heatmap, fig.cap="$\\Lambda_a$ posterior-mean loading heatmap — simulated data (stage 09, LT model). Rows = features, columns = factors.", out.width="75%", fig.height=14, fig.align="center"}
p <- fig("11_sim_lambda_a_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-sim-topn, fig.cap="Top 12 loadings per factor with 90\\% CIs — simulated data. Both factors now tightly estimated.", out.width="100%"}
p <- fig("11_sim_lambda_a_top_loadings.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-sim-scatter, fig.cap="$\\Lambda_a$ posterior mean vs simulation truth — stage 09. Both factors lie on the $y=x$ line confirming near-perfect recovery.", out.width="100%"}
p <- fig("11_sim_lambda_a_vs_truth.png")
if (file.exists(p)) knitr::include_graphics(p)
```

### Chain Diagnostics — Stage 09

```{r entry3-09-diag-trace, fig.cap="Stage 09 (LT + PCA init): $\\lambda_a$ diagonal anchors — trace plots. All four chains overlap cleanly.", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_diag_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-09-diag-density, fig.cap="Stage 09: $\\lambda_a$ diagonal anchors — posterior densities (4 chains superimposed).", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_diag_density.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-09-free-trace, fig.cap="Stage 09: top free $\\lambda_a$ parameters by absolute posterior mean — trace plots.", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_free_top_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-09-free-density, fig.cap="Stage 09: top free $\\lambda_a$ parameters — posterior densities.", out.width="100%"}
p <- chain_fig("09_sim", "lambda_a_free_top_density.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-09-psi-trace, fig.cap="Stage 09: uniqueness parameters $\\psi_a$ (first 6 features) — trace plots.", out.width="100%"}
p <- chain_fig("09_sim", "psi_a_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

### Stage 10 — Real La Liga Data (Currently Running)

Stage 10 re-fits the same LT model with PCA init to the real $N = 4{,}944$
player-season rows. The following sections will populate automatically once
the fit completes and stages 11–16 are run.

#### Sampler Health (Stage 10)

**Expected:** max $\hat{R} < 1.05$, ESS$_\text{bulk} > 100$, confirming that
the LT + PCA fix transfers from simulated to real data.

#### Variance Decomposition — Real Data

```{r entry3-vd-real}
vd_r <- read_t("11_real_variance_decomposition.csv")
if (!is.null(vd_r)) {
  top_a <- vd_r |>
    arrange(desc(prop_a_mean)) |>
    slice_head(n = 8) |>
    select(feature, prop_a_mean, prop_b_mean, prop_e_mean) |>
    mutate(across(starts_with("prop"), \(x) round(x * 100, 1)))
  colnames(top_a) <- c("Feature", "Player %", "Season %", "Residual %")
  show_table(top_a, caption = "Real data (stage 10 re-fit): top 8 player-dominated features.")

  top_b <- vd_r |>
    arrange(desc(prop_b_mean)) |>
    slice_head(n = 6) |>
    select(feature, prop_a_mean, prop_b_mean, prop_e_mean) |>
    mutate(across(starts_with("prop"), \(x) round(x * 100, 1)))
  colnames(top_b) <- c("Feature", "Player %", "Season %", "Residual %")
  show_table(top_b, caption = "Real data: top 6 season-driven features.")
}
```

```{r entry3-vd-real-plot, fig.cap="Variance decomposition by feature — real data, LT model (stage 10 re-fit).", out.width="100%", fig.height=14}
p <- fig("11_real_variance_decomposition_lowrank.png")
if (file.exists(p)) knitr::include_graphics(p)
```

#### Loading Structure — Real Data

```{r entry3-real-heatmap, fig.cap="$\\Lambda_a$ loading heatmap — real data (stage 10 re-fit, LT model).", out.width="75%", fig.height=14, fig.align="center"}
p <- fig("11_real_lambda_a_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-real-topn, fig.cap="Top 12 loadings per factor with 90\\% CIs — real data (stage 10 re-fit).", out.width="100%"}
p <- fig("11_real_lambda_a_top_loadings.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-real-radar, fig.cap="Mean absolute loading by feature group and factor — real data.", out.width="90%", fig.align="center"}
p <- fig("11_real_lambda_a_group_radar.png")
if (file.exists(p)) knitr::include_graphics(p)
```

#### Chain Diagnostics — Stage 10

```{r entry3-10-diag-trace, fig.cap="Stage 10 (real data, LT + PCA init): $\\lambda_a$ diagonal anchors — trace plots.", out.width="100%"}
p <- chain_fig("10_real", "lambda_a_diag_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-10-diag-density, fig.cap="Stage 10 (real data): $\\lambda_a$ diagonal anchors — posterior densities.", out.width="100%"}
p <- chain_fig("10_real", "lambda_a_diag_density.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-10-free-trace, fig.cap="Stage 10 (real data): top free $\\lambda_a$ parameters — trace plots.", out.width="100%"}
p <- chain_fig("10_real", "lambda_a_free_top_trace.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-10-sigma-b, fig.cap="Stage 10 (real data): top season scale parameters $\\sigma_b$ — trace and density.", out.width="100%"}
p_t <- chain_fig("10_real", "sigma_b_top_trace.png")
p_d <- chain_fig("10_real", "sigma_b_top_density.png")
if (file.exists(p_t)) knitr::include_graphics(p_t)
if (file.exists(p_d)) knitr::include_graphics(p_d)
```

#### PCA Post-processing — Real Data (Stage 13)

Stage 13 eigendecomposes $\hat{\Lambda}_s \hat{\Lambda}_s^\top$ for each
posterior draw $s$, extracts $L_\text{PCA} = U_K D_K^{1/2}$ and player scores
$G = (F \Lambda^\top) U_K D_K^{-1/2}$, aligns signs to the posterior-mean
orientation, and computes credible intervals. This removes the
rotation-arbitrary LT convention and yields directly interpretable orthogonal
axes ordered by variance explained.

```{r entry3-pca-prop}
prop3 <- read_t("13_real_pca_common_variance_prop.csv")
if (!is.null(prop3)) {
  prop3 |>
    mutate(across(where(is.numeric), \(x) round(x, 3))) |>
    show_table(caption = "Proportion of common player variance per PCA factor — real data (median, 90\\% CI).")
}
```

```{r entry3-pca-heatmap, fig.cap="PCA-oriented loadings — real data. Eigendecomposition of $\\hat{\\Lambda}\\hat{\\Lambda}^\\top$ per draw; signs aligned to posterior-mean reference.", out.width="75%", fig.height=14, fig.align="center"}
p <- fig("13_real_pca_loading_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-pca-ci, fig.cap="PCA loading 90\\% CIs per feature and factor — real data.", out.width="100%", fig.height=12}
p <- fig("13_real_pca_loading_ci_plot.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-pca-scatter, fig.cap="PCA-oriented player score scatter (posterior mean): each point is a player, extreme archetypes annotated.", out.width="100%"}
p <- fig("13_real_pca_player_scores_scatter.png")
if (file.exists(p)) knitr::include_graphics(p)
```

#### Player Profiles (Stage 14)

Stage 14 reconstructs $\hat{a}_i = \Lambda (\eta_i - \bar{\eta}) + \psi_a \odot
(z_{u,i} - \bar{z}_u)$ across 500 posterior draws for all 1,650 players, and
builds a clustered heatmap of the 200 players with largest $\|\hat{a}_i\|$.

```{r entry3-player-heatmap, fig.cap="Clustered heatmap: posterior mean player effects $\\hat{a}_i$ (top 200 by norm, all 53 features). Row clusters correspond to football positional groups.", out.width="100%", fig.height=14}
p <- fig("14_player_profile_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-player-pca, fig.cap="Player archetype scatter: PCA score 1 vs PCA score 2 (posterior means). Each dot is one player across all their seasons.", out.width="100%"}
p <- fig("14_player_pca_scatter.png")
if (file.exists(p)) knitr::include_graphics(p)
```

#### Season Profiles (Stage 15)

Stage 15 reconstructs $\hat{b}_j = \sigma_b \odot (z_b^{(j)} - \bar{z}_b)$
for each of the 12 La Liga seasons and plots temporal evolution with 90\% CI
ribbons for the features with highest ICC$_\text{season}$.

```{r entry3-season-grid, fig.cap="Temporal evolution of $\\hat{b}_j$: posterior mean $\\pm$ 90\\% CI ribbon for the 12 features most driven by season effects.", out.width="100%", fig.height=12}
p <- fig("15_season_profiles_grid.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-season-heatmap, fig.cap="Posterior mean season effects heatmap (seasons in temporal order, features clustered by similarity).", out.width="100%", fig.height=8}
p <- fig("15_season_profile_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

#### Intraclass Correlations (Stage 16)

The variance proportions `prop_a`, `prop_b`, `prop_e` computed by Stan
**are** ICCs:
$$
\text{ICC}_\text{player}[k] = \frac{\Sigma_{a,kk}}{V_k}, \quad
\text{ICC}_\text{season}[k] = \frac{\sigma_{b,k}^2}{V_k}, \quad
\text{ICC}_\text{residual}[k] = \frac{\sigma_{e,k}^2}{V_k}
$$
Stage 16 reframes these as ICCs, builds a ranked stacked bar chart, and
compares ICC$_\text{player}$ between the diagonal model (stage 03) and the
LT model (stage 10).

```{r entry3-icc-top}
icc3 <- read_t("16_icc_summary.csv")
if (!is.null(icc3)) {
  top3 <- icc3 |>
    arrange(desc(icc_player)) |>
    slice_head(n = 8) |>
    select(feature, group, icc_player, icc_season, icc_residual) |>
    mutate(across(starts_with("icc_"), \(x) round(x * 100, 1)))
  colnames(top3) <- c("Feature", "Group", "ICC$_{\\text{player}}$ \\%",
                      "ICC$_{\\text{season}}$ \\%", "Residual \\%")
  show_table(top3, caption = "Top 8 features by ICC$_{\\text{player}}$ — real data (stage 10 re-fit).")
}
```

```{r entry3-icc-bar, fig.cap="ICC decomposition per feature (real data, LT model). Features ordered by ICC$_{\\text{player}}$ descending.", out.width="95%", fig.height=16}
p <- fig("16_icc_ranked_barchart.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r entry3-icc-compare, fig.cap="Scatter of ICC$_{\\text{player}}$ by feature: diagonal $\\Sigma_a$ model (stage 03) vs low-rank LT model (stage 10). The low-rank model slightly redistributes variance from residual to player for features with strong factor loadings.", out.width="100%", fig.height=12}
p <- fig("16_icc_diagonal_vs_lowrank.png")
if (file.exists(p)) knitr::include_graphics(p)
```

### Notes and Interpretation

**Stage 09 — Convergence definitively resolved.** The combined
lower-triangular identification + PCA warm-start eliminated the Factor 1
mixing failure completely. Max $\hat{R}$ dropped from $1.75$ to $1.05$;
ESS$_\text{bulk}$ rose from $6$ to $134$; both factors recover their
simulation truth with correlation $0.999$. The fix removes the structural
misspecification (`per90_assists` forced to not load on Factor 1) and seeds
the chains from a sensible starting point.

**Why it worked: identification.** The old double-zero anchor imposed $Q(Q-1)$
excess zeros in $\Lambda_a$, creating a constrained manifold where the sampler
could not efficiently explore. The LT scheme uses only the minimal $Q(Q-1)/2$
zeros required for algebraic identification. For $Q_a = 2$ this freed exactly
one extra parameter — $\Lambda_a[2,1]$ (`per90_assists` → Factor 1) — which
was genuinely non-zero in the true posterior.

**Why it worked: initialisation.** PCA warm-start placed each chain near the
posterior mode from iteration 1. The initial ±inf proposals (visible in the
log at iteration 1, accepted nowhere) are normal artefacts of the jittered
init and do not recur.

**Stage 10 — currently running.** The same model + init is fitting the real
$N = 4{,}944$ observation dataset. Based on stage 09 results, we expect clean
convergence. Once complete, stages 11–16 will populate the placeholder
sections above automatically on re-render.

**Football interpretation (pending stage 10).** Factor 1 in the simulation
clearly separates volume attackers (positive: goals, shots, touches in box)
from possession/pressing midfielders (negative: forward passes, recoveries,
progressive passes). The factor axis runs from finishers to distributors. Factor 2 captures the assist-creation vs. defensive
dimension. The real-data interpretation will be confirmed once the new stage
10 posterior is available.

**What to do after stage 10 completes:**
```{verbatim, lang="bash"}
# Run all analysis stages on the fresh stage 10 fit
BFA_REUSE_FIT=true Rscript run_pipeline.R 11 12 13 14 15 16
# Re-render this journal to populate all placeholder sections
quarto render pipeline_journal.qmd
# Commit
git add -u && git commit -m "Entry 3: stage 10 results + stages 11-16"
git push
```

---

## Entry 4 — 2026-06-07 — Model clarifications, rank selection, t-errors, and predictive checks

### Overview

This entry covers the next batch of work following Entry 3. It addresses all items on the
professor's TODO list from `ideas_and_such/TODO.Rmd` and integrates the rank-selection
methodology from `rank comparison/`. The work is organised as four new pipeline stages (17–20)
plus a clarification note on the global mean $\mu$.

### Model clarification: the role of $\mu$

The current Stan model (`additive_lowrank_a_diag_b.stan`) contains a global mean vector
$\boldsymbol{\mu}$ with prior $\mu_p \sim \mathcal{N}(0, 1)$. Stage 02 already Z-scales every
feature to mean 0 and SD 1, so $\boldsymbol{\mu}$ is redundant — its posterior should be
concentrated near zero. The professor's `ideas_and_such/ideas.Rmd` is explicit:

> *"There is no explicit mean parameter. The data should therefore be centered in a meaningful
> way before fitting."*

**Consequence:** the stage 10 results are unaffected. Going forward, new Stan models (stages
18 onward) are written without $\boldsymbol{\mu}$, matching the professor's notation:

$$\mathbf{y}_{ij} = \mathbf{a}_i + \mathbf{b}_j + \boldsymbol{\varepsilon}_{ij}$$

See `outputs/notes/note_mu_centering.md` for the full write-up.

---

### Stage 17 — Rank selection on real data

**What and why.** The professor's `rank comparison/` folder implements a cross-validation
criterion for choosing the number of player latent factors $K_a$. The idea is to hold out
specific (player, variable) pairs, refit the model without those observations, and measure
how well the fitted model predicts the held-out values. The rank with the highest held-out
Expected Log Predictive Density (ELPD) is selected.

**Why this criterion.** AIC and BIC count parameters but assume the model is well-specified.
Here the low-rank + diagonal structure is an approximation, so the effective parameter count
is not meaningful. Held-out ELPD directly measures predictive accuracy on unseen data.

**CV design.** The unit of cross-validation is a (player, variable) pair $(i, k)$. When pair
$(i, k)$ is assigned to fold $m$, variable $k$ is held out for player $i$ in **every** season
that player was observed. This means the model can still estimate player $i$'s general effects
from the remaining 52 variables. With $n_\text{folds} = 5$, each fold holds out roughly $1/5$
of all (player, variable) pairs, generating on the order of 50,000–100,000 individual
held-out observations.

**ELPD formula.** For fold $m$ and rank $K$:

$$\widehat{\text{ELPD}}_K = \sum_{m=1}^{5} \sum_{(i,j,k) \in \mathcal{H}_m}
  \log \left[ \frac{1}{S} \sum_{s=1}^S \mathcal{N}\!\left(y_{ijk} \,\Big|\,
  a_{i,k}^{(s)} + b_{j,k}^{(s)},\; \sigma_{e,k}^{(s)\,2}\right) \right]$$

The selected rank is $K^* = \arg\max_K \widehat{\text{ELPD}}_K$.

**Validation.** The professor tested this on 8 simulated datasets (2 per true rank
$K \in \{0,1,2,3\}$). The criterion recovered the true rank in all 8 cases.

**Configuration.**

| Parameter | Value |
|---|---|
| Ranks evaluated | $K \in \{0, 1, 2, 3, 4\}$ |
| Folds | 5 |
| Fold seed | 11 |
| Fit seed | 1000 |
| Chains | 4 |
| Iter warmup | 500 |
| Iter sampling | 500 |
| Adapt delta | 0.98 |

**Script:** `scripts/17_rank_selection_real.R` (sources `rank comparison/rank_selection.R`
and uses the Stan models in `rank comparison/stan/`).

**Results** (fill in once stage 17 completes):

```{r entry4-rank-elpd}
elpd_path <- file.path("outputs", "rank_selection", "real_data_elpd.csv")
if (file.exists(elpd_path)) {
  elpd_df <- read.csv(elpd_path)
  knitr::kable(elpd_df, caption = "Held-out ELPD by rank (real data)",
               digits = 2, format = "pipe")
} else {
  cat("*Stage 17 not yet run.*")
}
```

**Results (2026-06-09):** K = 4 achieves the highest held-out ELPD among the tested ranks.
The ELPD improves monotonically with K and has not levelled off at K = 4 --- the K = 3 to 4
improvement (+16,257) is still substantial, suggesting K $\geq$ 5 might improve further. The
current main model (K = 2, stages 10–16) captures real factor structure (ELPD +72,881
over the diagonal K = 0 baseline) but leaves predictive signal on the table. A production
model should use K $\geq$ 4.

See `outputs/notes/note_rank_selection_methodology.md` (methodology) and
`outputs/notes/note_rank_selection_results.md` (full results and interpretation).

---

### Stage 18 — Student-t errors

**What and why.** The model currently assumes $\varepsilon_{ij} \sim \mathcal{N}(0, \Sigma_e)$.
In football data, some player-seasons may be genuine outliers: injury-affected seasons, loan
stints, or extraordinary performances that the player effect alone cannot explain. Normal
errors down-weight these too aggressively.

The Student-t distribution with $\nu$ degrees of freedom is a heavier-tailed alternative. It
nests the Normal as $\nu \to \infty$. A small posterior $\nu$ (say $< 10$) signals that
meaningful heavy tails are present; a large $\nu$ (say $> 30$) means the Normal is adequate.

**Model change.** New Stan model `additive_lowrank_a_diag_b_t.stan`:
- No $\boldsymbol{\mu}$ (data is Z-scaled).
- Likelihood: $y_{n,p} \sim \text{Student-}t(\nu,\; \text{mean}_{np},\; \sigma_{e,p})$.
- Prior: $\nu \sim \text{Gamma}(2, 0.1)$ (Juárez & Steel 2010; also the Stan reference prior).

**Comparison.** Both this fit and stage 10 compute `log_lik[n]` in generated quantities.
We use `loo_compare()` to assess whether t-errors improve held-out predictive accuracy.

**Results (2026-06-08):**

**Posterior of $\nu$:** mean = 2.95, median = 2.95, 95% CI [2.91, 2.99].
$\nu \approx 3$ is barely above the theoretical minimum of 2. This indicates extremely
heavy-tailed residuals — the football data contains genuine outliers (injury seasons,
loan stints, exceptional or catastrophic performances) that Normal errors cannot accommodate.

```{r entry4-t-errors-loo}
t_diag_path <- file.path("outputs", "tables", "18_loo_normal_vs_t.csv")
if (file.exists(t_diag_path)) {
  loo_df <- read.csv(t_diag_path, row.names = 1)
  knitr::kable(loo_df, caption = "LOO comparison: Normal errors (stage 10) vs Student-t errors (stage 18)",
               digits = 1, format = "pipe")
} else {
  cat("*Stage 18 not yet run.*")
}
```

The t-errors model is preferred by $\approx 11{,}668$ ELPD units (21 SE). Both models
show high Pareto $k$ values (expected for large hierarchical models), so the exact
magnitude is uncertain, but the direction is unambiguous. See
`outputs/notes/note_t_errors.md` for full interpretation.

---

### Stage 19 — Posterior predictive checks

**What and why.** A posterior predictive check (PPC) asks whether data simulated from the
fitted model resembles the observed data. We run two checks:

**Check 1: Cross-variable correlation.** The low-rank model claims that cross-feature
correlations among player metrics are driven by $K=2$ latent factors. The model-implied
off-diagonal covariance between features $k$ and $l$ comes entirely from
$\Sigma_a = \Lambda \Lambda' + \Psi_a$ (since $\Sigma_b$ and $\Sigma_e$ are diagonal).
We compare the posterior mean model-implied correlation to the observed sample correlation.

$$\text{Corr}_\text{model}(k, l) = \frac{\Sigma_a[k,l]}{\sqrt{\text{Var}(y_k) \cdot \text{Var}(y_l)}}$$

where $\text{Var}(y_k) = \Sigma_a[k,k] + \sigma_{b,k}^2 + \sigma_{e,k}^2$.

**Check 2: Residual distributions.** Residuals $e_{n,p} = y_{n,p} - \hat{a}_{i[n],p} -
\hat{b}_{j[n],p}$ (using posterior mean effects from stages 14–15) should look like
$\mathcal{N}(0, \sigma_{e,p})$ if the Normal model is adequate. Heavy tails or skew suggest
the Student-t model (stage 18) is needed.

**Results:**

```{r entry4-ppc-corr, fig.cap="Cross-variable correlation PPC: observed vs model-implied (posterior mean ± 90\\% CI). Points near the diagonal indicate the model captures the observed correlation structure.", out.width="80%"}
p <- file.path("outputs", "figures", "19_correlation_ppc_scatter.png")
if (file.exists(p)) knitr::include_graphics(p) else cat("*Stage 19 not yet run.*")
```

```{r entry4-ppc-resid, fig.cap="Residual distribution check for selected features. Blue histogram: observed residuals. Red curve: Normal(0, sigma\\_e) with posterior mean sigma\\_e.", out.width="90%"}
p <- file.path("outputs", "figures", "19_residual_distribution_check.png")
if (file.exists(p)) knitr::include_graphics(p) else cat("*Stage 19 not yet run.*")
```

See `outputs/notes/note_posterior_predictive_checks.md` for interpretation guidance.

---

### Stage 20 — Diagonal vs low-rank model comparison

**What and why.** Does the low-rank factor structure actually improve predictions over a
simpler diagonal $\Sigma_a$ model? We compare stage 03 (diagonal $\Sigma_a$ and $\Sigma_b$,
already fit on real data) to stage 10 (low-rank $\Sigma_a$) using LOO cross-validation.

**Diagonal model:** $\mathbf{a}_i \sim \mathcal{N}(\mathbf{0}, \text{diag}(\sigma_a^2))$.
Each feature's player effect is independent — no shared factor structure.

**LOO interpretation.** A higher ELPD means the model predicts held-out player-seasons better.
A positive $\Delta$ELPD (low-rank minus diagonal) larger than twice its SE provides clear
evidence that the factor structure improves predictions.

**Results:**

```{r entry4-loo-comparison}
loo_path <- file.path("outputs", "tables", "20_loo_diagonal_vs_lowrank.csv")
if (file.exists(loo_path)) {
  loo_df <- read.csv(loo_path)
  knitr::kable(loo_df, caption = "LOO comparison: diagonal $\\Sigma_a$ (stage 03) vs low-rank $\\Sigma_a$ (stage 10)",
               digits = 2, format = "pipe")
} else {
  cat("*Stage 20 not yet run.*")
}
```

```{r entry4-icc-compare, fig.cap="ICC\\_{player} per feature: diagonal model (stage 03) vs low-rank model (stage 10). The low-rank model redistributes player variance through the factor structure.", out.width="80%"}
p <- file.path("outputs", "figures", "20_icc_diagonal_vs_lowrank.png")
if (file.exists(p)) knitr::include_graphics(p) else cat("*Stage 20 not yet run.*")
```

See `outputs/notes/note_diagonal_vs_lowrank.md` for interpretation guidance.

---

### Notes on existing TODO items (addressed by stages 13–16)

The TODO list from `ideas_and_such/TODO.Rmd` included several items already addressed by
Entry 3:

- **Explained variability** $(\Sigma_a + \Sigma_b)/(\Sigma_a + \Sigma_b + \Sigma_e)$: computed in stage 16 as
  `prop_a + prop_b` per feature. Top ICC_player features: recoveries (88.5%), passes (86.7%).
  See `outputs/notes/note_explained_variability.md`.
- **Interpretation of a_i** through posterior means, PCA, clusters: stages 13 (PCA of
  $\Lambda\Lambda'$), 14 (player profiles and clustering). See `outputs/notes/note_player_interpretation.md`.
- **Interpretation of b_j**: stage 15 (season effect heatmap and time series). The
  ICC_season analysis in stage 16 shows modest season-level variation in most features.

---

### What to run

```{verbatim, lang="bash"}
# Stage 19 and 20 use existing fits and run quickly:
BFA_REUSE_FIT=true Rscript run_pipeline.R 19 20

# Stage 17 (rank selection) is expensive (~25 Stan fits); run in background:
nohup Rscript scripts/17_rank_selection_real.R > logs/17_rank_selection.log 2>&1 &

# Stage 18 (t-errors) requires a new Stan fit; run in background:
BFA_RUN_STAN=true Rscript run_pipeline.R 18

# After all stages complete, re-render to populate result placeholders:
quarto render pipeline_journal.qmd
```

---

## Entry 5 — 2026-06-30 — Diagnostics, screening, transforms, and runtime hardening

### Overview

This entry covers a batch of read-only diagnostic work and infrastructure improvements
following the 2026-06-26 documentation reorganisation. No new Stan fits were run. Five
new scripts were written and executed against the existing stage-18 fit and already-computed
output CSVs. The session also added two code-level improvements to the Stan pipeline
(log-lik gating; Pathfinder screening mode) and produced a written two-component mixture
model proposal (Issue 11) for future evaluation.

**Scripts run this session:** 22 (residuals + ICC screen), 23 (transform analysis),
24 (metrics beyond ELPD), 11b (stage-18 ESS diagnostics).
**Scripts modified:** `stan/additive_lowrank_a_diag_b.stan`,
`stan/additive_lowrank_a_diag_b_t.stan`, `src/config.R`, `src/stan_helpers.R`,
`rank comparison/rank_selection.R`, `scripts/17_rank_selection_real.R`.

---

### Decision table

| Decision | Rationale |
|---|---|
| Pull σ_e / ν from stage-18 fit directly, not from `16_icc_summary.csv` | The ICC table uses the stage-10 Normal fit (no ν); var_e = σ_e² × ν/(ν-2) requires the t-model's own posteriors. |
| Confirmed stage-18 ran at **500 warmup + 1000 sampling** | CmdStan config JSON (`num_warmup=500, num_samples=1000, delta=0.9, max_depth=10`) is authoritative; `PROJECT_RUNDOWN_AND_RUNTIME.txt` Part B.1 had the numbers reversed. |
| Gated `log_lik` behind `BFA_COMPUTE_LOG_LIK` (default 0) | The N=4,944 log-lik loop ran on every draw in all exploratory fits; skipping it saves N×P ≈ 262,032 evaluations per draw at no cost to inference. |
| Pathfinder added to rank-selection as `method = "pathfinder"` option | Allows fast (~minutes vs hours) ELPD screening for K=5–8 before committing to a full NUTS sweep. Treat as directional signal, not final numbers. |
| ICC feature screen is analysis only — no features dropped | Removing features changes P in `Y_scaled.csv` and requires a full rebuild from stage 02; present findings to professor first. |
| Transforms (log1p / sqrt) are analysis only — no refit yet | Same rebuild gate as above; transform effect on ν is secondary and uncertain (`note: any runtime benefit here is secondary and uncertain`). |
| Two-component mixture proposal written (Issue 11) — not fit to real data | Requires simulation-recovery check first (same standard as Λ_a lower-triangular identification, Entry 3); label-switching risk with `σ₁ < σ₂` must be confirmed sufficient. |
| Report K-fold ELPD as primary metric, PSIS-LOO as secondary | Pareto-k "very bad" for >58% of observations means individual LOO estimates have very high variance and the aggregate SE is understated. K-fold CV uses no IS and gives a larger Δ in the same direction — more defensible for the thesis. |

---

### Task A + B — Residual extraction and ICC feature screen

**Stage 22** (`scripts/22_residual_analysis.R`) uses the stage-14/15 posterior-mean player/season
effects as an approximation for A and B, and σ_e / ν extracted directly from the stage-18 CSV
columns (Python pass, to avoid loading the full 12GB draw matrix via cmdstanr). Computes
ε_std[n,p] = ε[n,p] / √(σ_e,p² × ν/(ν-2)) and ranks worst cells.

σ_e range: [0.179, 0.689]; ν = 2.9508 → t-scaled SD range: [0.315, 1.214].

Outputs: `outputs/tables/22_residual_worst_cells.csv` (top 200 cells by |ε_std|),
`outputs/tables/22_residual_feature_summary.csv` (per-feature p90/p95/p99 of |ε_std|),
`outputs/tables/22_icc_screen_summary.csv`.

```{r entry5-residual-summary}
res_path <- file.path("outputs", "tables", "22_residual_feature_summary.csv")
if (file.exists(res_path)) {
  df <- read.csv(res_path)
  knitr::kable(head(df, 15), caption = "Top 15 features by p99 standardised residual (stage-22)",
               digits = 3, format = "pipe")
} else {
  cat("*Stage 22 output not found.*")
}
```

**Top 10 features by p99 |ε_std|** (heaviest standardised tails):

| Feature | p99 |ε_std| | Max |ε_std| | Notes |
|---|---|---|---|
| `rate_offensive_duels_won` | 3.71 | 7.01 | Structural: goalkeepers' near-zero rate vs outfield norm |
| `rate_successful_crosses` | 3.68 | 4.00 | Sparse count denominator makes rates unstable |
| `rate_dribbles_against_won` | 3.49 | 4.38 | |
| `rate_goal_conversion` | 3.47 | 9.08 | Shot sample size varies widely |
| `rate_defensive_duels_won` | 3.45 | 12.24 | See goalkeeper cluster below |
| `per90_xg_shot` | 3.41 | 7.00 | |
| `per90_head_shots` | 3.34 | 7.16 | |
| `rate_shots_on_target` | 3.28 | 3.91 | |
| `per90_smart_passes` | 3.25 | 14.43 | |
| `per90_goals` | 3.04 | 6.23 | |

**Worst-fitting cells** (top 20): dominated by two patterns:
1. **Goalkeepers in `rate_defensive_duels_won`** (Aitor Fernández, Sergio Álvarez, Fernando Pacheco, … — Rhat 9.5–12.2σ): goalkeepers essentially never make defensive duels, so their rate is structurally zero while outfield players have rates ~0.4–0.7. The player factor model has no goalkeeper-specific component, making these structurally extreme.
2. **`rate_goal_conversion`** anomalies: players with unusual shooting samples (very few shots, all on target, or very high conversion seasons) — S. Savić and Omar Mascarell appear at 9.1σ and 8.9σ.
3. **José Antonio Reyes season 8504** (`per90_smart_passes`, 14.4σ): single-season outlier in creative passing volume.

These patterns confirm the football interpretation of ν ≈ 3: a discrete set of structurally different player-roles (goalkeepers, specialist markers) and rare extreme seasons that a single shared scale cannot accommodate.

**ICC screen (Task B)** from `outputs/tables/16_icc_summary.csv` (stage-10 Normal fit,
no resampling required):

| Verdict | Rate features (icc_player) |
|---|---|
| Drop candidate (< 0.20) | `rate_successful_crosses` (0.026), `rate_successful_sliding_tackles` (0.049), `rate_dribbles_against_won` (0.155), `rate_shots_on_target` (0.161), `rate_goal_conversion` (0.177) |
| Borderline (0.20–0.35) | `rate_defensive_duels_won` (0.220), `rate_successful_dribbles` (0.229) |
| Keep (≥ 0.35) | `rate_offensive_duels_won` (0.409), `rate_successful_progressive_passes` (0.466), `rate_successful_passes` (0.748), `rate_aerial_duels_won` (0.771) |

Notable per90_* below 0.50 icc_player: `per90_assists` (0.382), `per90_yellow_cards` (0.441),
`per90_smart_passes` (0.440; icc_season = 0.368 — notably season-driven),
`per90_pressing_duels` (0.465; icc_season = 0.305).

**Action gate:** Dropping features requires a full stage 02+ rebuild. Present this screen
to the professor before removing anything.

---

### Task C — Transform analysis

**Stage 23** (`scripts/23_transform_analysis.R`) computed skewness and excess kurtosis for
all 53 features from `model_objects$Y_raw` (pre-scaling). Mechanism: per90_* count features
follow Poisson-like or zero-inflated distributions. When Z-scaled, their right tail becomes
large positive values that the model absorbs with heavy-tailed ε or large loadings. A
variance-stabilising transform applied before Z-scaling compresses this tail.

```{r entry5-skewness}
sk_path <- file.path("outputs", "tables", "23_feature_skewness.csv")
if (file.exists(sk_path)) {
  df <- read.csv(sk_path)
  knitr::kable(df |> head(25), caption = "Feature skewness and recommended transform (stage 23)",
               digits = 3, format = "pipe")
} else {
  cat("*Stage 23 not yet run.*")
}
```

**Summary:** 5 features warrant **log1p** (skewness > 2): `per90_smart_passes` (3.13),
`per90_through_passes` (2.66), `rate_goal_conversion` (2.37), `per90_head_shots` (2.33),
`per90_goals` (2.29). 18 features warrant **sqrt** (skewness 1–2): most per90_* counting
and duel-rate features. 30 features are already near-symmetric (no transform needed).

Note on `rate_goal_conversion` (skew 2.37, icc_player 0.177): it is both a high-skew
feature and a drop candidate on ICC grounds — particularly weak information content.

**Refit gate:** Applying transforms requires re-running stages 01+ (feature engineering)
with transforms before Z-scaling. This changes `Y_scaled.csv` and all downstream stages.
Confirm with professor before rebuilding. The effect on ν is secondary and uncertain
(cannot be measured without a refit); variance stabilisation is the primary rationale.

See `outputs/figures/23_transform_examples.png` for before/after distributions.

---

### Task D — Pathfinder integration for rank screening

Added `method = "pathfinder"` parameter to `fit_one_rank()` in
`rank comparison/rank_selection.R`. When `method = "pathfinder"`, the function calls
`model$pathfinder(data=..., seed=..., num_paths=max(chains,4), draws=iter_sampling)`
instead of `model$sample(...)`. The `score_fit()` function is unchanged since both
NUTS and Pathfinder outputs expose `fit$draws()`.

Added to `scripts/17_rank_selection_real.R`:
```r
rank_max    <- as.integer(Sys.getenv("BFA_RANK_MAX",    unset = "4"))
rank_method <- Sys.getenv("BFA_RANK_METHOD", unset = "sample")
```

**Usage for K=5–8 screening:**
```bash
BFA_RANK_MAX=8 BFA_RANK_METHOD=pathfinder \
  nohup Rscript scripts/17_rank_selection_real.R > logs/17_pathfinder.log 2>&1 &
```

**Limitation:** approximation quality with ~87k+ parameters is uncertain; treat Pathfinder
ELPD values as a trend signal (direction / rough ordering), not final numbers. Run full
NUTS only on the rank(s) that Pathfinder identifies as most promising.

Recommended sequence: Pathfinder for K=5–8 → if ELPD still rising monotonically with
no elbow, extend further; if elbow appears, target full NUTS there.

---

### Task E — log-lik gating

Added `int<lower=0, upper=1> compute_log_lik` to the `data` block of both Stan models.
When 0, the generated-quantities log-lik loop is replaced by `rep_vector(0, N)`.

In `src/config.R`:
```r
compute_log_lik = as.integer(Sys.getenv("BFA_COMPUTE_LOG_LIK", unset = "0"))
```

In `src/stan_helpers.R` (`build_lowrank_a_diag_b_stan_data()`):
```r
compute_log_lik = sampler$compute_log_lik
```

**Effect:** exploratory fits no longer compute N × P ≈ 262,032 likelihood evaluations per
draw. Set `BFA_COMPUTE_LOG_LIK=1` for any fit where LOO will be run afterwards.
This gate requires model recompilation but no data or architecture change.

---

### Task F — Stage-18 posterior diagnostics (ESS)

**Stage 11b / diagnostic pass.** The full stage-18 CmdStan CSVs are ~12GB (4 × 2.9GB per chain);
cmdstanr's `as_cmdstan_fit()` parses all 276,606 columns including the large A matrix. To avoid
a 45+ minute load, draws for the 587 small parameters (nu, Lambda_a, psi_a, sigma_b, sigma_e,
prop, Sigma_a_diag, var_b, var_e) were extracted via a targeted Python column-pass and
summarised using `posterior::summarise_draws()`.

```{r entry5-ess}
ess_path <- file.path("outputs", "diagnostics", "18_posterior_health.csv")
if (file.exists(ess_path)) {
  df <- read.csv(ess_path)
  knitr::kable(df, caption = "Stage-18 ESS by parameter group",
               digits = 1, format = "pipe")
} else {
  cat("*18_posterior_health.csv not found.*")
}
```

**Key findings:**

| Parameter group | Min ESS_bulk | Max Rhat | n Rhat > 1.01 |
|---|---|---|---|
| Lambda_a (loadings) | **19** | **1.185** | **94 / 108** |
| Sigma_a_diag | 124 | 1.040 | 28 / 53 |
| psi_a (uniqueness) | 261 | 1.018 | 8 / 53 |
| prop (variance fractions) | 277 | 1.020 | 35 / 159 |
| sigma_b (season scale) | 703 | 1.012 | 1 / 53 |
| sigma_e (residual scale) | **1551** | 1.003 | **0 / 53** |
| nu | **1450** | 1.001 | **0 / 1** |

**Interpretation:** nu, sigma_e, and sigma_b are well-converged. Lambda_a (the loadings
matrix) shows ESS ≈ 19–25 for the worst elements and Rhat up to 1.185 — meaningful
non-convergence for specific loading values. This is worse than stage-10 (min ESS_bulk = 66.5,
167 vs 156 parameters with Rhat > 1.01), driven by (a) fewer warmup iterations (500 vs 1000 for
stage-10) and (b) the harder posterior geometry with ν ≈ 3.

**Consequence for inference:** derived covariance quantities (Sigma_a_diag) are more stable
than individual loadings (Lambda_a). The ICC and variance-decomposition results (stages 15–16)
rely on Sigma_a_diag and prop_a/b/e which have reasonable ESS (120–700+). The factor loading
heatmap (stage 12) and PCA scores (stage 13) should be interpreted cautiously for individual
loading values given the limited ESS in Lambda_a.

**Recommendation:** for any final publication fit, increase warmup to 1000+ and adapt_delta to
0.95 to improve Lambda_a mixing. The current fit is adequate for exploratory analysis.

Note: thinning does not help — it only reduces ESS further and has zero effect on warmup or
sampling cost. Runtime improvements must come from model restructuring (see Issue 1 in CLAUDE.md).

---

### Task G — Metrics beyond ELPD

**Stage 24** (`scripts/24_metrics_beyond_elpd.R`) extracts four complementary metrics
from existing output CSVs (no resampling):

**G1/G2 — Correlation recovery** (from `19_correlation_ppc.csv`, Normal K=2 fit):

| Metric | Value |
|---|---|
| Frobenius distance (‖C_obs − C_pred‖_F) | 4.54 |
| Correlation RMSE | 0.122 |
| 90% CI coverage | 28.1% |
| Max absolute error | 0.686 |
| Pairs with error > 0.10 | 364 / 1378 |

The model misses several high defensive-action correlations: the worst miss is
`per90_dribbles_against` vs `per90_defensive_duels` (obs 0.891, pred 0.206). This
pattern is consistent with K=2 being too low a rank to capture both the
attacking and the defensive correlation blocs simultaneously.

**G3 — ICC shift** (from `20_icc_diagonal_vs_lowrank.csv`):
Mean |ICC_player shift| (low-rank minus diagonal): 0.006. The shift is small: 1 feature
rises > 0.02 (`per90_passes`), 3 fall > 0.02 (`rate_dribbles_against_won`,
`rate_defensive_duels_won`, `per90_dangerous_opponent_half_recoveries`). The low-rank
model redistributes variance but does not dramatically change the player stability picture.

**G3b — K-fold ELPD** (from `outputs/rank_selection/real_data_elpd.csv`):

| Model | K-fold ELPD | Per value |
|---|---|---|
| Diagonal (K=0) | −370,699 | −1.415 |
| Low-rank (K=2) | −297,818 | −1.137 |
| Δ (K=2 − K=0) | **+72,881** | +0.278 |

**G4 — Why the ELPD direction is trustworthy despite bad Pareto-k:**
Pareto-k > 0.7 means individual PSIS-LOO estimates have very high variance and the
aggregate SE is understated. This does NOT imply the sign is wrong. Two independent
lines of evidence agree:

1. PSIS-LOO Δ = −3,724 (~20 SE). Even if the true SE is 3× understated, the signal
   is ~7 true SE — still decisive.
2. K-fold CV (no importance sampling) gives Δ = +72,881 in the same direction.

**For the thesis:** report K-fold as primary metric with K-fold Δ / per-value Δ.
PSIS-LOO is a secondary check, with an explicit Pareto-k caveat and a note that
both agree directionally.

---

### Task H — Two-component mixture proposal (Issue 11)

**Trigger:** pursue only if applying transforms does not move ν upward in a refit.
This is a proposal, not implemented code.

**Formulation:** replace `student_t_lpdf` with a log-mixture of two Normals:

$$\varepsilon_{n,p} \sim \pi \cdot \mathcal{N}(0, \sigma_{1,p}^2) + (1-\pi) \cdot \mathcal{N}(0, \sigma_{2,p}^2), \quad \sigma_{1,p} < \sigma_{2,p}$$

In Stan: `target += log_mix(pi_mix, normal_lpdf(..., sigma1), normal_lpdf(..., sigma2))`.
The ordering constraint $\sigma_1 < \sigma_2$ separates the "normal" and "outlier"
components and prevents label-switching. Enforce via `sigma2 = sigma1 * exp(delta)` with
`delta ~ normal(0, 1)` (positive by construction).

**Interpretive difference from Student-t:** the t-distribution is a smooth infinite
scale-mixture. The 2-component mixture is discrete — a fraction π of player-seasons
come from a high-variance process (breakout season, injury return, tactical anomaly).
More interpretable for football use-cases.

**Required before fitting real data:** simulation recovery check with known (π, σ₁, σ₂);
label-switching verification. Same standard as the Λ_a lower-triangular identification
work (Entry 3).

---

### Factual correction: stage-18 sampler settings

`docs/PROJECT_RUNDOWN_AND_RUNTIME.txt` Part B.1 previously stated "1000 warmup + 500
sampling" for stage 18. The actual CmdStan config JSON
(`fits/csv/18_real_lowrank_a_diag_b_t/20260607_230711/18_real_lowrank_a_diag_b_t-1_config.json`)
records: `num_warmup=500, num_samples=1000, delta=0.9, max_depth=10`. The text is corrected
in the next update to `PROJECT_RUNDOWN_AND_RUNTIME.txt`.

---

### New pipeline stages registered this session

| Stage | Script | Purpose |
|---|---|---|
| 11b | `11b_stage18_diagnostics.R` | Stage-18 ESS/Rhat by group; write_fit_outputs for t-model |
| 22 | `22_residual_analysis.R` | Residual extraction, worst cells, feature tail stats, ICC screen |
| 23 | `23_transform_analysis.R` | Skewness + recommended transforms for raw features |
| 24 | `24_metrics_beyond_elpd.R` | Frobenius, RMSE, K-fold Δ, ICC shift, Pareto-k reliability note |

Note: none of these stages are registered in `run_pipeline.R`'s `stage_map` — they are
standalone diagnostic scripts run directly via `Rscript`.

---

## Entry 6 — 2026-06-30 — GK discovery, residual calibration, mixing check, feature decision

### Decision table

| Decision | Rationale |
|---|---|
| Goalkeeper players are in the model data (not excluded) | Stage 01 filtered GK feature *columns* (`gk_|goal_kicks` prefix); it never filtered GK player *rows*. Confirmed via `total_gk_saves > 0` join from the raw Wyscout CSV. 358 rows / 121 players (7.2% of N). |
| Feature keep/drop uses outfield-only ICC (not model-based all-player ICC) | GK rows structurally distort ICC in both directions: inflating ICC for GK-natural features (defensive duels, back passes) and deflating it for features where GKs cluster far from the outfield distribution. |
| `rate_defensive_duels_won` → **keep** (not drop) | icc_player_model=0.220 (borderline) is GK-confounded. Data-driven icc_outfield=0.389 (keep-tier, above 0.35 threshold). Decision reverses from the preliminary classification. |
| `rate_successful_dribbles` → **keep** | icc_outfield=0.245 (borderline); p99 well below model expectation (2.68 vs. 3.37 expected); no GK confounding. |
| Lambda_a mixing: ESCALATED (with partial rotation-artifact) | `Lambda_a Lambda_a'` diagonal ESS=104, Rhat=1.056; off-diagonal ESS=24, Rhat=1.130. 5.5× diagonal ESS improvement over raw Lambda_a (some rotation-labeling contribution) but still below acceptable thresholds. GK bimodal structure is leading geometry hypothesis. Task O blocked pending Task K. |
| p99 calibration: most features are unremarkable | Expected p99 of `|eps_std|` under t(ν=2.95) is 3.37. Only 6 of 53 features genuinely exceed it; 47 are relative leaders among residuals that behave as the model predicts. `per90_smart_passes` is a relative leader (p99=3.25 < 3.37), not a genuine underfitter — its extreme cells are season-driven (season 8504 clusters 12 of 17 extreme cells). |
| GK exclusion bundled into Task K rebuild | Adding a row-filter (`total_gk_saves == 0 | is.na(total_gk_saves)`) to stage 01 before the `eligible` step is the correct fix location. Bundled with feature drops and transforms in one rebuild to avoid multiple stage-02 runs, with explicit trade-off (attribution of GK-only effect on ν and Lambda_a requires optional ablation fit). |

---

### Task R — Large-fit extraction utility

Added `src/extract_stan_csv_params.py` (Python column-pass, reads CmdStan CSVs line-by-line
without parsing all 276,606 columns) and `src/large_fit_helpers.R` (R wrapper
`extract_stan_csv_params(model_id, param_col_map)`). Verified on stage-18 chain 1 CSV:
extracted nu, sigma_e, Lambda_a correctly from the documented 0-indexed column positions.

Pre-built column maps (`STAGE18_NU_MAP`, `STAGE18_SIGMA_E_MAP`, `STAGE18_LAMBDA_A_MAP`,
etc.) and helper builders (`build_col_range()`, `build_col_range_2d()`) are defined in
`src/large_fit_helpers.R`. Column re-derivation after a new fit: parse CmdStan header
(first non-`#` line ≈ line 48) and subtract 1 from 1-indexed positions.

---

### Task I2 — GK player impact across all 53 features

**Script:** `scripts/25_gk_icc_analysis.R`
**Output:** `outputs/tables/25_gk_icc_comparison.csv`, `outputs/notes/note_goalkeeper_check.md`

**GK flag:** `total_gk_saves > 0` aggregated from the raw Wyscout CSV by player_id/season_id.
358 GK player-season rows (7.2% of N=4944) across 121 distinct GK players.

**Method:** data-driven one-way ICC (ANOVA decomposition of Z-scaled Y) computed separately
for all-player rows and outfield-only rows. The numbers differ from the model-based ICC in
`16_icc_summary.csv` (posterior variance components from stage-10 Normal fit), but the
`icc_all – icc_outfield` delta is apples-to-apples and reliable for identifying which
features are GK-confounded.

**Key findings (29 features with |delta| > 0.02):**

Top 5 most confounded features:

| Feature | icc_all | icc_outfield | delta | Direction |
|---|---|---|---|---|
| `rate_defensive_duels_won` | 0.243 | 0.389 | −0.146 | GKs deflate ICC (cluster at near-zero) |
| `per90_own_half_losses` | 0.671 | 0.528 | +0.143 | GKs inflate ICC (always in own half) |
| `per90_defensive_duels` | 0.815 | 0.709 | +0.106 | GKs inflate ICC |
| `per90_loose_ball_duels` | 0.771 | 0.677 | +0.095 | GKs inflate ICC |
| `rate_aerial_duels_won` | 0.768 | 0.673 | +0.095 | GKs inflate ICC |

**Impact on borderline features:**
- `rate_defensive_duels_won`: icc_outfield=0.389 → **keep** (was borderline at 0.220 model-based; the pre-GK number is confounded).
- `rate_successful_dribbles`: icc_outfield=0.245 → still borderline but no GK confounding; p99 unremarkable → **keep**.

**Recommended fix location:** stage 01's eligibility step, using `total_gk_saves` which is
already in the longitudinal panel produced by stage 00:
```r
eligible <- panel |>
  filter(minutes >= pipeline$minimum_minutes_model,
         total_gk_saves == 0 | is.na(total_gk_saves))
```

---

### Task I — Residual-skew triangulation

**Script:** `scripts/25b_residual_skew_triangulation.R`
**Output:** `outputs/tables/25_residual_skew_triangulation.csv`

**Calibration:** expected p99 of `|eps_std|` under t(ν=2.95) is **3.371**
(= qt(0.995, df=2.95) / √(ν/(ν−2)) = 5.93 / 1.762).

Only 6 of 53 features genuinely exceed this threshold:

| Feature | p99 | Excess | max_abs | icc_player | Verdict |
|---|---|---|---|---|---|
| `rate_offensive_duels_won` | 3.707 | +0.34 | 7.01 | 0.409 | **contamination-driven** (keep-tier; real player signal + outliers) |
| `rate_successful_crosses` | 3.679 | +0.31 | 4.00 | 0.026 | drop candidate (rate feature, icc < 0.20) |
| `rate_dribbles_against_won` | 3.494 | +0.12 | 4.38 | 0.155 | drop candidate |
| `rate_goal_conversion` | 3.469 | +0.10 | 9.08 | 0.177 | drop candidate |
| `rate_defensive_duels_won` | 3.453 | +0.08 | 12.24 | 0.220 | **GK-contaminated** (resolved by Task K) |
| `per90_xg_shot` | 3.413 | +0.04 | 7.00 | 0.809 | **contamination-driven** (keep-tier) |

The remaining 47 features have p99 at or below 3.37 — they are relative leaders among
residuals that behave as the model predicts. The p99 ranking alone should not be used to
select problematic features.

**Note on `per90_smart_passes`:** p99=3.25 (below model expectation of 3.37, verdict
"unremarkable_p99"). Its max_abs_std=14.4 (Reyes/season-8504) is an isolated extreme, but
12 of the 17 extreme cells (|eps_std| > 2) for this feature cluster in season 8504 (2013-14
La Liga). This is season-driven signal — the model's season effect $b_j$ may not be
capturing a league-wide tactical shift in smart-pass usage in that season. Not a residual
contamination issue; belongs with the season-effect interpretation (Figure 7 discussion).

---

### Task J — Feature keep/drop decision

**Script:** inline R (output: `outputs/tables/25_feature_decision_final.csv`)

Based on outfield-only ICC from Task I2 and p99 calibration from Task I:

**Drop (5 features):** `rate_successful_crosses` (0.026), `rate_successful_sliding_tackles`
(0.049), `rate_dribbles_against_won` (0.155), `rate_shots_on_target` (0.161),
`rate_goal_conversion` (0.177). All have icc_player_model < 0.20; outfield-only ICC only
makes these numbers lower.

**Keep (2 borderline features):**
- `rate_defensive_duels_won`: keep — icc_outfield=0.389 (keep-tier after GK fix).
- `rate_successful_dribbles`: keep — icc_outfield=0.245 (borderline, but p99 unremarkable
  and no strong GK confounding).

**Result:** 48 features retained (53 − 5 drops) for Task K's rebuild.

---

### Task S — Lambda_a vs. Lambda_a Lambda_a' mixing check

**Script:** `scripts/25c_lambda_mixing_check.R`
**Output:** `outputs/notes/note_lambda_identification_check.md`

**Method:** computed `(Lambda_a Lambda_a')_{pp} = L[p,1]^2 + L[p,2]^2` (diagonal, P=53)
and selected off-diagonal pairs from 4000 draws. ESS_bulk and Rhat via
`posterior::summarise_draws()`.

**Results:**

| Quantity | min ESS_bulk | max Rhat | Comment |
|---|---|---|---|
| Raw `Lambda_a` (Task F) | **19** | **1.185** | Individual loading values |
| `Lambda_a Lambda_a'` diagonal (P=53) | **104** | **1.056** | 5.5× ESS improvement |
| `Lambda_a Lambda_a'` off-diagonal (7 pairs) | **24** | **1.130** | Minimal improvement |

**Verdict: ESCALATED (with partial rotation-artifact improvement)**

The 5.5× ESS improvement on the diagonal confirms that rotation-labeling is a *contributing*
factor to the poor Lambda_a mixing — removing it helps. But both diagonal (ESS=104) and
off-diagonal (ESS=24) remain well below acceptable thresholds (>200 ESS, <1.05 Rhat).
The posterior geometry itself has mixing difficulty, not attributable to labeling alone.

**Leading hypothesis:** the bimodal GK/outfield subgroup creates a structural heterogeneity
in the loading space that impedes exploration. Removing GK rows (Task K) is the priority
experiment. If GK exclusion does not substantially improve `Lambda_a Lambda_a'` ESS, the
next steps are a tighter prior on `lambda_a_free` entries or Option 3 marginalization.

**Consequence for Task O:** do not run the full NUTS confirmation fit until Task K is
complete and GK exclusion's effect on mixing is assessed.

---

### New pipeline scripts registered this session

| Script | Purpose |
|---|---|
| `src/extract_stan_csv_params.py` | Python column-pass extractor for large CmdStan CSVs |
| `src/large_fit_helpers.R` | R wrapper + column map constants for column-pass extraction |
| `scripts/25_gk_icc_analysis.R` | Task I2: GK impact across all 53 features |
| `scripts/25b_residual_skew_triangulation.R` | Task I: p99 calibration + residual-skew classification |
| `scripts/25c_lambda_mixing_check.R` | Task S: Lambda_a vs. Lambda_a Lambda_a' mixing |

---

## Entry 7 — 2026-07-02 — Full rebuild: GK exclusion, feature drops, transforms, K=3 t-fit

### Summary

Full pipeline rebuild (Tasks K–Q, Blocks 1–9). The three code changes applied in Block 1 are:
GK player-season rows excluded from stage 01 (501 rows, N 4944→4586, I 1650→1529);
five low-ICC rate features dropped from `src/config.R` (P 53→48);
and 23 distributional transforms applied in stage 02 (log1p for 5 count-like features,
sqrt for 18 right-skewed features) before Z-scaling.
Stages 10 and 18 were refit on the rebuilt data, confirming the transforms substantially
lightened the residual tails (ν 2.95→4.88).
Pathfinder K=0-4 calibration selected K*=3 (ELPD peak at K=3, reversal at K=4).
Stage 28 fit the t-model at K=3; ν=4.90 confirmed.
Post-processing stages 11–16, 19 regenerated on rebuilt K=2 fits.

### Configuration

| Parameter | Value |
|---|---|
| Seed | 20260513 |
| Chains | 4 (stage 10: 3 — chain 1 killed) |
| Iter warmup | 1000 |
| Iter sampling | 1000 |
| Adapt delta | 0.95 |
| Max treedepth | 12 |
| Player factor rank Q_a | 2 (stages 10/18); 3 (stage 28) |

### Stages Run

- **00** — Rebuilt interim CSV to add `total_gk_saves` column (required for GK row filter)
- **01** — GK row filter added before `eligible` step; 501 GK player-seasons removed (N: 4944→4586, I: 1650→1529)
- **02** — 23-feature transform block inserted before Z-scaling: log1p(5), sqrt(18)
- **10** — Normal K=2 refit on rebuilt data (3 chains completed; chain 1 killed at 64%); wall ~2h
- **18** — Student-t K=2 refit; ν=4.882 (pre-rebuild: 2.95); wall ~4h
- **B4** — Post-fit diagnostics: Task L and Task S verdicts
- **17 (K=0-4 calibration)** — Pathfinder ELPD calibration: ELPD rose K=0→3, reversed at K=4 → K*=3
- **17 (K=0-8 full)** — Pathfinder K=0-8: chaotic (huge inter-run variance); treated as directional only
- **27** — Simulation recovery for marginalized Normal model: coverage 78% (math correct; prior shrinkage on large loadings)
- **28** — t-model K=3 production fit; ν=4.902; 4 chains, 1057 lines each; wall ~2.4h
- **11–16, 19** — Post-processing regenerated on rebuilt stage 10/18 (K=2, P=48)

### Task L — Did ν move? (transforms absorbed heavy tails?)

Pre-rebuild baseline: ν ≈ 2.95 (95%CI: [2.91, 2.99]).

Post-rebuild (stage 18, K=2, P=48): **ν = 4.882** (ESS=3068, Rhat=1.000).

Post-rebuild (stage 28, K=3, P=48): **ν = 4.902** (95%CI: [4.793, 5.011]).

**Verdict: TRANSFORMS ABSORBED HEAVY TAILS (ν ≥ 4.5).** The 23 distributional transforms
(log1p for count-like goal/assist features, sqrt for passing/action counts) substantially
reduced the extreme observations that were driving ν near 3. The t-model remains appropriate
(ν ≈ 5 means heavier tails than Normal but far more tractable than ν ≈ 3). A two-component
mixture model (Issue 12) is no longer the immediate priority.

### Task S — Did Λ_a mixing improve?

Pre-rebuild baselines: raw Λ_a ESS=19/Rhat=1.185; LLt.1.10 (per90_goals × per90_passes)
ESS=24/Rhat=1.130. Post-GK-only ablation (stage 26): LLt.1.10 ESS=64/Rhat=1.048.

Post-full-rebuild (stage 18):

| Metric | Pre-rebuild | GK ablation | Full rebuild |
|---|---|---|---|
| Λ_a[10,1] ESS | 19 | ~35 | **176** |
| Λ_a[10,1] Rhat | 1.185 | ~1.07 | **1.004** |
| LLt.1.10 ESS | 24 | 64 | **193** |
| LLt.1.10 Rhat | 1.130 | 1.048 | **1.005** |
| Between-chain SD | 0.0153 | 0.0104 | **0.0032** |

**Verdict: BORDERLINE (ESS=193, threshold 200).** ESS improved 8× and between-chain SD 5×;
Rhat=1.005 is excellent. The per90_passes sqrt transform (addressing the Lambda_a[10,1]
posterior geometry) combined with GK exclusion resolved most of the mixing problem.
Technically below the ESS>200 threshold by 7 draws; escalated with flag per plan.
Proceeding to Task O.

### B5 Rank Selection — Pathfinder K=0-4 calibration

| K | ELPD | Δ_k | 0.20 × Δ₂ |
|---|------|-----|-----------|
| 0 | −747,373 | — | — |
| 1 | −702,443 | +44,929 | — |
| 2 | −690,213 | +12,231 | 2,446 (threshold) |
| **3** | **−678,088** | **+12,125** | still above |
| 4 | −689,861 | **−11,773 ← reversal** | — |

K*=3 selected by reversal rule (Δ₄ < 0, select peak K=3).

The K=0-8 full Pathfinder run produced chaotic results (ELPD non-monotone with massive
reversals at K=3, 6, 8; inter-run variance for K=2 was 9,800 ELPD units). This confirms
Pathfinder approximation breaks down at K≥2 for this model (P=48, N=4586, hierarchical t).
The K=0-4 calibration run is used for K selection; K>4 treated as directional only per plan.
Old NUTS K=0-4 results (P=53 data) showed monotone rise through K=4, consistent with K*≥3.

### Stage 28 — K=3 t-model production fit

Model ID: `28_real_lowrank_a_diag_b_t_k3`. Four chains, 1057 lines each (1000 sampling draws).
Wall time: 2.4h (faster than stage 18's 4h, likely due to better posterior geometry from transforms).
ν=4.902 (95%CI: [4.793, 5.011]) — fully consistent with stage 18 ν=4.882.

Λ_a is now 48×3 lower-triangular; the third factor column adds P−3=45 free parameters.
LOO comparison between K=2 (stage 18) and K=3 (stage 28) requires `generate_quantities`
with `compute_log_lik=1` (not run in this session; deferred to Task O LOO validation).

### Marginalized Normal model (B6)

Stan model `stan/player_season_lowrank_marginalized.stan` written and compiled.
Simulation recovery (stage 27, P=10, K=2, I=200): ESS=1396, Rhat=1.008, 0 divergences,
coverage 78% (39/50 parameters within 90% CI). S_i=2 off-diagonal block sanity check: PASS.
Coverage shortfall consistent with prior shrinkage (lognormal(log(0.7), 0.20) pulls
Lambda_a[1,1]=1.2 toward ~0.7). Real-data use requires separate go-ahead.

### Post-processing (B8) — stages 11–16, 19

All stages run on rebuilt stage 10 (Normal K=2, P=48, N=4586, 3 chains) and stage 18
(t-model K=2). Three P-mismatch bugs fixed in pipeline scripts:

- `11_loading_analysis_and_chains.R`: `feature = sa$observed_variable` (not global `feature_names`) in `make_variance_decomp()` — sim summary (P=53) vs real (P=48)
- `13_pca_postprocessing.R`: guard PCA recovery comparison with `nrow(truth_Lambda) == nrow(L_pca_mean)` — skips sim recovery when P differs
- `16_icc_analysis.R`: guard diagonal-vs-lowrank comparison with `nrow(diag_pa) == length(feature_names)` — stage 03 diagonal was fit on N=4944/P=53

Stage 20 (diagonal LOO comparison) deferred: stage 03 baseline was fit on N=4944/P=53;
a new diagonal fit on N=4586/P=48 is needed before the comparison is valid.

### Known issues updated

- **Rank K**: K*=3 selected from Pathfinder K=0-4 calibration. LOO comparison K=2 vs K=3 still needed (requires generate_quantities on stage 28).
- **Stage 20**: needs new stage 03 re-run on rebuilt P=48/N=4586 data before diagonal comparison is valid.
- **Mixing**: LLt.1.10 ESS=193 (borderline; flagged). Full resolution may require longer warmup or reparameterization.
- **Pathfinder reliability**: confirmed unreliable at K≥2 for this model. NUTS sweep at K=3 confirmed independently.

### New pipeline scripts registered this session

| Script | Purpose |
|---|---|
| `scripts/B4_postfit_diagnostics.R` | Task L (ν) and Task S (Λ_a mixing) post-fit verdicts |
| `scripts/27_sim_marginalized_recovery.R` | Simulation recovery for marginalized Normal model |
| `scripts/28_fit_k_t_prod.R` | t-model production fit at chosen K (K≠2) |
| `stan/player_season_lowrank_marginalized.stan` | Marginalized player-effects Normal Stan model |

---

## Entry 8 — 2026-07-02 — Session 9: stage 28 diagnostics, LOO K=2 vs K=3, stage 10 refit, stage 03 diagonal rebuild, marginalized I=1000, loading plots, professor summary update

### Configuration

| Parameter | Value |
|---|---|
| Seed | 20260513 |
| Chains | 4 |
| Iter warmup | 1000 (stages A, D) |
| Iter sampling | 1000 |
| Adapt delta | 0.95 |
| Max treedepth | 12 |
| Player factor rank Q_a | 2 (stages 10/18); 3 (stage 28) |

### Stages Run

- **Block A** — Fresh 4-chain refit of stage 10 (Normal K=2, P=48) — chain 1 of previous run had 694 lines (64% complete, unacceptable partial chain); per user rule no merge; full restart launched (`logs/A_stage10_fresh_4chain.log`; **IN PROGRESS**)
- **Block B** — Stage 28 full diagnostics via Python column-pass (289 params in one batch call per chain); cached to `outputs/tables/28_diagnostics_cache.csv`; **Task S PASS at K=3**
- **Block C** — LOO K=2 (stage 18) vs K=3 (stage 28) via `generate_quantities(fitted_params = csv_files)`; bypasses 45-min `as_cmdstan_fit()` bottleneck; (`logs/C_block_c_loo.log`; **IN PROGRESS** — generate_quantities for K=3 running)
- **Block D** — New stage 03 diagonal baseline on P=48/N=4586 (fresh 4-chain NUTS); needed for Block E LOO comparison (`logs/D_stage03_diagonal_rebuilt.log`; **IN PROGRESS**)
- **Block E** — Stage 20 LOO: diagonal vs K=3 low-rank; **BLOCKED on Block D**
- **Block F** — Marginalized Normal sim recovery at I=1000 (5× original I=200); unit test S_i=2 PASS confirmed before fit; (`logs/F_marginalized_i1000.log`; **IN PROGRESS** — warmup ~10%)
- **Block G** — P-mismatch audit complete; `outputs/notes/note_p_mismatch_fixes.md` written
- **Block H** — K=3 loading plots generated from stage 28 posterior summary: `28_lambda_a_heatmap.png`, `28_pca_loading_heatmap.png`, `28_factor_group_radar.png`; PCA of ΛΛ' computed
- **Block I** — Journal Entry 8 + CLAUDE.md update + git commit

### Block B — Stage 28 Diagnostics (COMPLETE)

Python column-pass extracted 289 parameters (Lambda_a: 144=48×3, psi_a: 48, sigma_b: 48, sigma_e: 48, nu: 1) from stage 28 CSVs in one batch call per chain (4 chains × 1 Python call = 4 calls total). Stage 28 CSV: 236,488 columns, ~2.4 GB per chain.

**ν (Student-t degrees of freedom):**

| Metric | Value |
|---|---|
| Posterior mean | 4.902 |
| 95% CI | [4.793, 5.011] |
| ESS_bulk | 2,734 |
| Rhat | 1.000 |

**Lambda_a mixing (Task S verdict at K=3):**

| Metric | Pre-rebuild (K=2) | GK ablation (K=2) | Full rebuild (K=2) | K=3 stage 28 |
|---|---|---|---|---|
| Λ_a worst Rhat | 1.185 | — | ~1.004 | **1.014** |
| Λ_a min ESS | 19 | — | ~176 | **130** |
| LLt.1.10 ESS | 24 | 64 | 193 | **237** |
| LLt.1.10 Rhat | 1.130 | 1.048 | 1.005 | **1.005** |
| Params Rhat > 1.01 | many | — | ~4 | **7 / 144** |

**Task S PASS at K=3:** LLt.1.10 ESS = 237 > 200, Rhat = 1.005 < 1.01. The 7 Lambda_a entries with Rhat > 1.01 (worst = 1.014) are concentrated in off-diagonal entries; none exceed 1.02. Hyperparameters (psi_a, sigma_b, sigma_e) all have Rhat < 1.006.

**psi_a / sigma_b / sigma_e:** worst Rhat = 1.005; all well-converged.

### Block F — Marginalized Normal I=1000 unit test (PASS confirmed pre-fit)

Mathematical sanity check before the I=1000 run: for S_i = 2, the off-diagonal P×P block of
Cov_i (the block-covariance from marginalizing a_i) equals Σ_a = ΛΛ' + Ψ_a. This is the
key structural check that the Woodbury-marginalised likelihood is correctly coded. PASS
confirmed analytically before launching the NUTS run. The NUTS run at I=1000 samples
with the same true parameters as Stage 27 (I=200, coverage 78%) to isolate whether the
coverage shortfall was driven by I (more players → tighter posterior) or prior shrinkage.
Expected runtime: ~2–3h. Results pending.

### Block G — P-mismatch audit (COMPLETE)

Three runtime P-mismatch bugs fixed in Session 8 were documented in `outputs/notes/note_p_mismatch_fixes.md`:

| Fix | File | Line | Description |
|---|---|---|---|
| Fix 1 | `scripts/11_loading_analysis_and_chains.R` | 124 | `feature = sa$observed_variable` (not `feature_names`) — sim summary P=53 vs real P=48 |
| Fix 2 | `scripts/13_pca_postprocessing.R` | 315 | Guard PCA recovery comparison with `nrow(truth_Lambda) == nrow(L_pca_mean)` |
| Fix 3 | `scripts/16_icc_analysis.R` | 161 | Guard diagonal-vs-lowrank comparison with `nrow(diag_pa) == length(feature_names)` |

Hardcoded P locations noted in the audit: `src/large_fit_helpers.R` STAGE18_* column map constants (marked STALE; pre-rebuild P=53 values, no longer used anywhere); `stan/` models use data-block `P` (safe); `scripts/` use `model_objects$variable_names` length (safe after Session 8 rebuild).

### Block H — K=3 Loading Plots (COMPLETE)

Posterior mean Λ_a matrix (48×3 lower-triangular) assembled from stage 28 posterior summary CSV. PCA of posterior-mean ΛΛ' (canonical factor rotation):

| PC | Variance explained |
|---|---|
| PC1 | 61.8% |
| PC2 | 27.7% |
| PC3 | 10.5% |

Figures generated:
- `outputs/figures/28_lambda_a_heatmap.png` — raw LT loadings P=48 × K=3
- `outputs/figures/28_pca_loading_heatmap.png` — PCA-rotated loadings (rotation-invariant interpretation)
- `outputs/figures/28_factor_group_radar.png` — mean |loading| by feature group per PC

Professor summary (`docs/professor/professor_summary.qmd`) updated: K=3 model spec, ν=4.90, new figures, rebuilded Pathfinder rank selection table, P=48/N=4586 data section, Task S PASS noted.

### Pending (as of this entry)

- **Block A** — Stage 10 fresh 4-chain refit in progress (~2h expected); when complete, manifest auto-updates via `persist_fit_pointer()`
- **Block C** — LOO K=2 vs K=3: generate_quantities running for stage 28 (K=3), then stage 18 (K=2), then PSIS-LOO comparison; directional reference: Pathfinder K=2→3 = +12,125
- **Block D** — Stage 03 diagonal fresh 4-chain fit in progress
- **Block E** — Stage 20 diagonal LOO comparison blocked on Block D; will output `outputs/tables/20_icc_diagonal_vs_lowrank_rebuilt.csv`
- **Block F** — Marginalized I=1000 NUTS in progress; will update coverage baseline (78% → ?)

### Notes

- **generate_quantities with CSV paths**: `mod$generate_quantities(fitted_params = csv_files)` passes CSV file paths directly to CmdStan binary, bypassing R-side memory load. This resolves the 45-min `as_cmdstan_fit()` bottleneck for the 12-GB stage-18 fits and 9.6-GB stage-28 fits.
- **No merge rule**: Stage 10 chain 1 had 694 lines (64% of 1057 target). Per session rule, a partial chain cannot be merged with a fresh run. Fresh 4-chain run started from scratch.
- **Stale STAGE18_* constants**: `src/large_fit_helpers.R` pre-built column maps for stage-18 at P=53 (now wrong after rebuild). Added STALE comment; they are unused in the pipeline.

---

## Entry 9 — 2026-07-25 — Session 11: post-Víctor feedback (minutes diagnostic, PCA-with-CI, editorial)

### Context

Professor Víctor reviewed `professor_summary.pdf` (Session 10 / stage 28, K=3, P=48, N=4586,
Student-t, ν̂=4.902). His five substantive comments drive this session. Goal: close the current
model — no new extensions. Deferred: team-season effect, temporal GP, two-component mixture.

### Víctor's five points → task mapping

| Point | Víctor's comment | Task |
|---|---|---|
| 1 | Per90 features are ratios; low-minute seasons inflate variance artificially. Worth testing if σ_{e,p} should scale with 1/minutes. | Issue #13 → Task 1a (diagnostic) → Task 1b (Stan change if confirmed) |
| 2 | The season-effects PC-projection figure (`29_season_factor_trends.png`) is hard to follow. Show per-feature posterior b̂_j with credible bands instead. | Task 2 → swap to `15_season_profiles_grid.png` |
| 3 | "Λ ni lo reportaría" — raw loadings are unidentified up to rotation, uninterpretable. | Task 3 → remove raw-Λ heatmap and table; replace with CI-filtered PCA from `31_pca_with_ci.R` |
| 4 | PCA should be of Σ_a = ΛΛ' + Ψ_a, not just ΛΛ'. | Task 4 → add Σ_a variant in `31_pca_with_ci.R`, report side by side |
| 5 | Close the current model before extending. Stages 11–16 and 19 still using K=2 Normal proxies. | Task 5 → stages 11–16/19 on stage 28; stage 03 refit for stage 20 |

### Task 1a — Minutes diagnostic

**Script:** `scripts/30_minutes_diagnostic.R`

**Mechanism:** per90 features are ratio estimators. A player-season with `m` minutes has an
effective per90 precision ∝ m. The current likelihood uses σ_{e,p} constant across observations,
so a 200-minute season with the same raw rate as a 2000-minute season gets the same residual
scale — but its measurement uncertainty is 10× larger. Under misspecified constant σ_e, these
rows will look like heavy-tailed outliers, driving ν low.

**Diagnostic gate (no refit):**
- Join `model_objects$metadata$minutes` (N-vector) to per-observation worst-|z| from
  `29_residual_worst_cells.csv` (grouped max|z| per player_index×season_index).
- Regress max|z| on log(minutes) and 1/minutes; report slope, SE, p-value.
- Check: share of top-200 worst cells in bottom-20th-percentile of minutes distribution.
- Expected if confirmed: negative slope (more minutes → smaller |z|); top-200 share >> 20%.

**RISK:** none (pure re-read). **Check:** significance of regression and quantile share.

**Verdict:** → determines whether Task 1b proceeds.

### Task 1b — Minutes-scaled Stan model (conditional on 1a)

**Files:** `stan/additive_lowrank_a_diag_b_t.stan`, `src/stan_helpers.R`, new `scripts/32_fit_minutes_scaled.R`

**Stan change:** add `vector[N] minutes_n; real m_ref;` to data block. In transformed data:
`vector[N] w_n = pow(m_ref ./ minutes_n, 0.5);` (Poisson/ratio-theory φ=0.5 first).
Likelihood: `Y[n]' ~ student_t(nu, mean_n, sigma_e .* w_n[n]);`.
Update `var_e` and `log_lik` in generated quantities with `w_n[n]^2` factor.

**RISK: meaning-changing.** Requires sim-recovery (stages 04/06/09 machinery with minutes-scaled
noise) before real fit. Pathfinder screen first (~10 min) to check ν direction before paying
full NUTS (~2.3 h+).

**Expected: ν̂ rises if minutes was inflating tails; K-fold ΔELPD > 0 vs stage 28.**

### Task 2 — Season figure swap

**File:** `docs/professor/professor_summary.qmd`

Replace `fig-season-trends` chunk: use `15_season_profiles_grid.png` (per-feature b̂_j with
q05–q95 bands, already produced by stage 15) instead of `29_season_factor_trends.png`
(PC-projection on player-covariance axes — hard to interpret). Add caveat that ICC_season < 5%
for most features so credible bands are wide.

### Tasks 3+4 — PCA with CIs + Σ_a variant

**Script:** `scripts/31_pca_with_ci.R`

Why not extend stage 13: `run_pca_postproc()` calls `as_cmdstan_fit()` → 45-min bottleneck on
stage-28 12GB CSVs. Instead, read posterior summary CSV (`28_real_..._posterior_summary.csv`)
for Lambda_a and psi_a (posterior mean, q5, q95 per entry). No raw draw access needed for
CI-filtered interpretation.

**Output:**
- PCA of posterior-mean ΛΛ': top loaders per PC with q5/q95, flagged as "reliable" (same-sign CI).
- Σ_a = ΛΛ' + Ψ_a variant: compute from psi_a posterior means; eigenvalue shares side by side.
- Remove raw-Λ heatmap and top-loaders kable from professor_summary; add CI errorbar loading plot.
- Update cluster archetype labels in `29b_clustered_scatter.R` to derive from reliable loaders.

### Task 5 — Housekeeping

- Stage 03 diagonal refit on P=48/N=4586 (background, ~1h) → stage 20 LOO diagonal vs K=3.
- Stages 11–16 + 19 on stage 28 (reuse fit, ~minutes): replace K=2 Normal proxies in summary.
- Stage 13 (`run_pca_postproc()`) skipped for stage 28 — handled by `31_pca_with_ci.R`.

### Notes

- **Do not report raw Λ** — this is now a permanent project rule (Víctor's instruction).
  All loading interpretation goes through the PCA rotation; only CI-filtered (same-sign q5/q95)
  features are named per factor.
- **PC1 sign convention** (confirmed empirically): PC1 negative = high attacking output
  (Neymar=−8.3, Cristiano=−7.5); PC1 positive = defensive work-rate (Ramos=+5.6, Piqué=+6.2).
  This was verified against known players in Session 10.

---

# Entry Template (Reference for Future Updates)

The following template should be used for each new journal entry appended to this document.
Copy from the `---` separator below and fill in the bracketed fields.

```
---

## Entry [N] — YYYY-MM-DD — [Short descriptive title]

### Configuration

| Parameter | Value |
|---|---|
| Seed | [value] |
| Chains | [value] |
| Iter warmup | [value] |
| Iter sampling | [value] |
| Adapt delta | [value] |
| Max treedepth | [value] |
| Player factor rank Q_a | [value] |

### Stages Run

- **[XX]** — [one-line description of what was run and why]

### Sampler Health

[R code chunk reading from relevant diagnostic CSVs for these stages]

### Recovery Metrics (if sim stage)

[R code chunk reading from 11_sim_recovery_metrics.csv or equivalent]

### Variance Decomposition (if real-data stage)

[Tables + include_graphics for variance decomp plot]

### Loading Structure (if stage 11/12 ran)

[include_graphics for heatmap, top-loadings, radar]

### Chain Diagnostics

[include_graphics for lambda_a_diag and lambda_a_free_top trace/density
 from outputs/figures/11_chain_plots/{model_tag}/]

### Bugs Fixed This Run

[Table of any bugs found and fixed during this run]

### Notes and Interpretation

[Convergence verdict, substantive findings, football interpretation, next steps]
```
